---
title: "Third South African National Burden of Disease Study"
subtitle: "Comprehensive analytical report: mortality estimates for South Africa, 1997–2019"
author: "NBD3 analytical team"
date: "`r format(Sys.Date(), '%d %B %Y')`"
runtime: shiny
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: textmate
    fig_caption: true
    css: www/report.css
    df_print: default
    self_contained: true
---
```{r setup, include=FALSE}
options(
  stringsAsFactors = FALSE,
  scipen = 999,
  shiny.sanitize.errors = TRUE
)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_packages <- c(
  "arrow", "cachem", "data.table", "dplyr", "highcharter", "htmltools",
  "knitr", "rmarkdown", "scales", "shiny", "yaml"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop(
    "Missing report package(s): ", paste(missing_packages, collapse = ", "),
    ". Run source('install_packages.R') from the repository root."
  )
}

find_report_root <- function(
    start = dirname(normalizePath(knitr::current_input(), mustWork = TRUE))) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(current, "_targets.R")) &&
        file.exists(file.path(current, "report", "R", "report_data.R"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) stop("Could not locate the NBD repository root.")
    current <- parent
  }
}

root <- find_report_root()
knitr::opts_knit$set(root.dir = root)
source(file.path(root, "report", "R", "report_data.R"))
source(file.path(root, "report", "R", "report_cache.R"))
source(file.path(root, "report", "R", "report_charts.R"))

# Small result slices are safe to cache at application scope. This lets later
# sessions on the same Shiny worker reuse selections calculated by earlier
# sessions without loading the analytical draw archive into memory.
report_cache_mb <- suppressWarnings(as.numeric(
  Sys.getenv("NBD3_REPORT_CACHE_MB", "96")
))
if (!is.finite(report_cache_mb) || report_cache_mb < 32) report_cache_mb <- 96
shiny::shinyOptions(
  cache = cachem::cache_mem(max_size = report_cache_mb * 1024^2)
)

config <- read_viz_config(root)
runtime <- load_viz_runtime_data(config, open_rates = TRUE)
ui_cache_version_key <- if (
    !is.null(runtime$full_uncertainty) &&
      identical(runtime$full_uncertainty$mode, "summary_cache")) {
  paste(
    runtime$full_uncertainty$manifest$cache_version %||% "cache",
    runtime$full_uncertainty$manifest$built_at %||% "unknown",
    runtime$full_uncertainty$n_draws,
    sep = "|"
  )
} else {
  "dynamic-or-point-only"
}
comparison_data <- data.table::as.data.table(data.table::copy(runtime$comparisons))
cause_uncertainty <- data.table::as.data.table(data.table::copy(
  runtime$cause_uncertainty %||% data.table::data.table()
))
catalog <- runtime$catalogs
injury_inputs <- load_injury_report_inputs(config)
injury_broad <- summarise_injury_broad_fractions(injury_inputs)
injury_broad_comparison <- data.table::as.data.table(data.table::copy(
  injury_broad$comparison
))
injury_detailed_comparison <- data.table::as.data.table(data.table::copy(
  injury_broad$detailed_comparison
))
injury_level_audit_path <- report_method_asset(
  config,
  "03_injury_survey_reference_audit.csv",
  fallback = file.path(
    root, "output", "tables", "03_injury_survey_reference_audit.csv"
  )
)
injury_level_audit <- if (file.exists(injury_level_audit_path)) {
  data.table::fread(injury_level_audit_path)
} else {
  data.table::data.table()
}

uncertainty_draw_n <- if (!is.null(runtime$full_uncertainty)) {
  as.integer(runtime$full_uncertainty$n_draws)
} else if (nrow(cause_uncertainty) && "n_draws" %in% names(cause_uncertainty)) {
  values <- cause_uncertainty$n_draws[is.finite(cause_uncertainty$n_draws)]
  if (length(values)) max(values) else NA_integer_
} else {
  NA_integer_
}
uncertainty_draw_label <- if (length(uncertainty_draw_n) == 1L && !is.na(uncertainty_draw_n) && is.finite(uncertainty_draw_n)) {
  paste0(format(uncertainty_draw_n, big.mark = ",", scientific = FALSE), " joint draws")
} else {
  "joint draws"
}
model_names_available <- ordered_model_names(unique(comparison_data$model), config)

measure_choices <- c(
  "Deaths" = "deaths",
  "Crude mortality rate per 100,000" = "crude_rate",
  "Age-standardised rate per 100,000" = "asr"
)

format_value <- function(x, measure, digits = 1L) {
  if (!length(x) || is.na(x) || !is.finite(x)) return("—")
  if (identical(measure, "fraction")) {
    return(scales::percent(x, accuracy = 0.1))
  }
  format(round(x, digits), big.mark = ",", scientific = FALSE, trim = TRUE)
}

metric_decimals <- function(measure) {
  if (identical(measure, "deaths")) 0 else 2
}

safe_default <- function(values, preferred = NULL) {
  values <- as.character(values)
  values <- values[!is.na(values) & nzchar(values)]
  if (!length(values)) return(character())
  if (!is.null(preferred) && as.character(preferred) %in% values) {
    return(as.character(preferred))
  }
  values[[1L]]
}


nbd_cached_reactive <- function(
    key_reactive,
    value_function,
    millis = 200L) {
  debounced_key <- shiny::debounce(key_reactive, millis = as.integer(millis))
  value <- shiny::reactive({
    value_function(debounced_key())
  })
  shiny::bindCache(
    value,
    debounced_key(),
    ui_cache_version_key,
    cache = "app"
  )
}

named_choices <- function(data, value, label) {
  data <- data.table::as.data.table(data.table::copy(data))
  if (!all(c(value, label) %in% names(data))) return(stats::setNames(character(), character()))
  data <- unique(data[, c(value, label), with = FALSE])
  keep <- !is.na(data[[value]]) & !is.na(data[[label]]) &
    nzchar(as.character(data[[value]])) & nzchar(as.character(data[[label]]))
  data <- data[keep]
  stats::setNames(as.character(data[[value]]), as.character(data[[label]]))
}

humanise <- function(x) tools::toTitleCase(gsub("_", " ", as.character(x)))
format_count <- function(x) {
  if (!length(x) || is.na(x) || !is.finite(x)) return("—")
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_table_value <- function(x, measure = NULL, digits = 1L) {
  vapply(
    x,
    function(value) {
      if (is.na(value) || !is.finite(value)) return("—")
      if (!is.null(measure) && identical(measure, "fraction")) {
        return(scales::percent(value, accuracy = 0.1))
      }
      format(
        round(value, digits),
        big.mark = ",",
        scientific = FALSE,
        trim = TRUE,
        nsmall = if (digits > 0L) digits else 0L
      )
    },
    character(1)
  )
}

nbd_kable_html <- function(
    data,
    caption = NULL,
    align = NULL,
    escape = TRUE,
    table_class = "nbd-kable",
    wrap_class = "nbd-kable-wrap",
    empty_message = "No data are available for this selection.") {
  x <- as.data.frame(data, stringsAsFactors = FALSE, check.names = FALSE)
  if (!nrow(x)) {
    x <- data.frame(Note = empty_message, check.names = FALSE)
  }
  if (is.null(align)) {
    align <- rep("l", ncol(x))
  }
  table_html <- knitr::kable(
    x,
    format = "html",
    escape = escape,
    align = align,
    caption = caption,
    table.attr = paste0('class="', table_class, '"')
  )
  htmltools::HTML(paste0(
    '<div class="', wrap_class, '">',
    as.character(table_html),
    '</div>'
  ))
}

comparison_panel_ui <- function(prefix, domain_name, default_series) {
  available <- comparison_data[get("domain") == domain_name]
  geographies <- unique(available[, .(
    geography_type, geography_code, geography
  )])[order(geography_type, geography_code)]
  sexes <- unique(available[, .(sex_code, sex)])[order(sex_code)]
  ages <- unique(available[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  series <- unique(available[, .(
    series_id, series_label, series_sort_order
  )])[order(series_sort_order)]

  ids <- paste0(prefix, c("_geography", "_sex", "_age", "_series"))
  shiny::tagList(
    shiny::fluidRow(
      shiny::column(
        3,
        htmltools::div(
          class = "control-panel",
          shiny::selectizeInput(
            ids[[1L]], "Geography",
            choices = named_choices(geographies, "geography_code", "geography"),
            selected = safe_default(geographies$geography_code, 10L),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[2L]], "Sex",
            choices = named_choices(sexes, "sex_code", "sex"),
            selected = safe_default(sexes$sex_code, 3L),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[3L]], "Age group",
            choices = named_choices(ages, "age_id", "age_label"),
            selected = safe_default(ages$age_id, "age_all"),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[4L]], "Series",
            choices = named_choices(series, "series_id", "series_label"),
            selected = safe_default(series$series_id, default_series),
            multiple = FALSE
          ),
          htmltools::p(
            class = "control-hint",
            "All available models are shown together. Click a legend item to hide or restore a series."
          )
        )
      ),
      shiny::column(
        9,
        htmltools::div(
          class = "report-chart",
          highcharter::highchartOutput(paste0(prefix, "_chart"), height = "500px")
        )
      )
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the annual model values"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download comparison data",
          class = "download-link"
        )
      )
    )
  )
}

register_comparison_panel <- function(prefix, domain_name, default_series) {
  ids <- list(
    geography = paste0(prefix, "_geography"),
    sex = paste0(prefix, "_sex"),
    age = paste0(prefix, "_age"),
    series = paste0(prefix, "_series"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  selection_key <- shiny::reactive({
    shiny::req(
      input[[ids$geography]], input[[ids$sex]],
      input[[ids$age]], input[[ids$series]]
    )
    list(
      panel = prefix,
      geography = as.integer(input[[ids$geography]]),
      sex = as.integer(input[[ids$sex]]),
      age = as.character(input[[ids$age]]),
      series = as.character(input[[ids$series]])
    )
  })

  selected_data <- nbd_cached_reactive(
    selection_key,
    function(key) {
      d <- comparison_data[
        get("domain") == domain_name &
          geography_code == key$geography &
          sex_code == key$sex &
          age_id == key$age &
          series_id == key$series
      ]
      nbd_rows <- d[model == "NBD3-R"]
      if (nrow(nbd_rows) && !is.null(runtime$full_uncertainty)) {
        dynamic <- tryCatch(
          collect_full_comparison_uncertainty(runtime, nbd_rows),
          error = function(error) {
            warning(
              "Model-comparison uncertainty could not be calculated: ",
              conditionMessage(error),
              call. = FALSE
            )
            data.table::data.table()
          }
        )
        if (nrow(dynamic)) {
          join_key <- c(
            "geography_type", "geography_code", "sex_code", "year",
            "age_id", "series_id", "measure"
          )
          dynamic <- dynamic[, c(join_key, "lower", "upper"), with = FALSE]
          static_external <- d[model != "NBD3-R"]
          revised_nbd <- merge(
            nbd_rows[, c("lower", "upper") := NULL],
            dynamic,
            by = join_key,
            all.x = TRUE,
            sort = FALSE
          )
          d <- data.table::rbindlist(
            list(static_external, revised_nbd),
            use.names = TRUE,
            fill = TRUE
          )
        }
      }
      d[order(
        year,
        factor(
          model,
          levels = as.character(unlist(
            config$labels$models$order,
            use.names = FALSE
          ))
        )
      )]
    },
    millis = 120L
  )

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 500))
    selected_measure <- unique(d$measure)[[1L]]
    d[, `:=`(
      plot_estimate = estimate,
      plot_lower = lower,
      plot_upper = upper
    )]
    suffix <- ""
    decimals <- metric_decimals(selected_measure)
    y_title <- measure_label(selected_measure, config)
    if (identical(selected_measure, "fraction")) {
      d[, `:=`(
        plot_estimate = 100 * plot_estimate,
        plot_lower = 100 * plot_lower,
        plot_upper = 100 * plot_upper
      )]
      suffix <- "%"
      decimals <- 1
      y_title <- "Percent of all-cause deaths"
    }
    hc_nbd_lines(
      d,
      y = "plot_estimate",
      series = "model",
      lower = "plot_lower",
      upper = "plot_upper",
      title = unique(d$series_label)[[1L]],
      subtitle = paste(unique(d$geography), unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = y_title,
      config = config,
      decimals = decimals,
      suffix = suffix,
      height = 500,
      filename = paste0(prefix, "-model-comparison"),
      series_order = as.character(unlist(config$labels$models$order, use.names = FALSE))
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No model series are available for the selected cell."
      ))
    }
    model_names <- ordered_model_names(unique(d$model), config)
    interval_models <- ordered_model_names(
      unique(d[is.finite(lower) & is.finite(upper), model]),
      config
    )
    interval_text <- if (length(interval_models)) {
      paste0(
        "Uncertainty bands are available for: ",
        paste(interval_models, collapse = ", "), "."
      )
    } else {
      "No uncertainty limits are supplied for this exact selection."
    }
    htmltools::div(
      class = "narrative-box",
      htmltools::strong("Models shown: "),
      paste(model_names, collapse = ", "),
      htmltools::br(),
      interval_text,
      htmltools::br(),
      htmltools::span(
        class = "provisional-note",
        paste0(
          "NBD3-R intervals come from ", uncertainty_draw_label,
          "; external intervals are retained as supplied."
        )
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    selected_measure <- unique(d$measure)[[1L]]
    digits <- if (identical(selected_measure, "fraction")) {
      1L
    } else {
      metric_decimals(selected_measure)
    }
    display <- d[, .(
      Model = model,
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Series = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      Source = source
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, selected_measure, digits),
      Lower = format_table_value(Lower, selected_measure, digits),
      Upper = format_table_value(Upper, selected_measure, digits)
    )]
    nbd_kable_html(
      display,
      caption = "Annual estimates and uncertainty limits for the selected model comparison",
      align = c("l", "l", "l", "r", "l", "l", "l", "r", "r", "r", "l")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-model-comparison.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )

  invisible(TRUE)
}

final_panel_ui <- function(prefix, geography_type_name, domain_name, default_series) {
  availability <- unique(catalog$availability[
    model == "NBD3-R" &
      get("geography_type") %in% geography_type_name &
      get("domain") == domain_name
  ])
  geographies <- unique(availability[, .(
    geography_code, geography
  )])[order(geography_code)]
  sexes <- unique(availability[, .(sex_code, sex)])[order(sex_code)]
  ages <- unique(availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  causes <- unique(availability[, .(
    series_id, series_label, hierarchy, series_sort_order
  )])[order(series_sort_order)]
  grouped_causes <- lapply(
    split(causes, causes$hierarchy),
    function(x) named_choices(x, "series_id", "series_label")
  )
  names(grouped_causes) <- humanise(names(grouped_causes))

  ids <- paste0(prefix, c(
    "_geographies", "_sex", "_cause", "_age", "_measure", "_years"
  ))
  shiny::tagList(
    htmltools::div(
      class = "control-panel final-control-panel",
      shiny::fluidRow(
        shiny::column(4, shiny::selectizeInput(
          ids[[1L]], "Geographies",
          choices = named_choices(geographies, "geography_code", "geography"),
          selected = as.character(geographies$geography_code),
          multiple = TRUE
        )),
        shiny::column(2, shiny::selectizeInput(
          ids[[2L]], "Sex",
          choices = named_choices(sexes, "sex_code", "sex"),
          selected = safe_default(sexes$sex_code, 3L),
          multiple = FALSE
        )),
        shiny::column(6, shiny::selectizeInput(
          ids[[3L]], "Cause",
          choices = grouped_causes,
          selected = safe_default(causes$series_id, default_series),
          multiple = FALSE
        ))
      ),
      shiny::fluidRow(
        shiny::column(4, shiny::selectizeInput(
          ids[[4L]], "Age group",
          choices = named_choices(ages[age_id != "asr_all"], "age_id", "age_label"),
          selected = safe_default(ages$age_id, "age_all"),
          multiple = FALSE
        )),
        shiny::column(4, shiny::selectInput(
          ids[[5L]], "Measure",
          choices = measure_choices,
          selected = "crude_rate"
        )),
        shiny::column(4, shiny::sliderInput(
          ids[[6L]], "Years",
          min = as.integer(config$labels$build$start_year),
          max = as.integer(config$labels$build$end_year),
          value = c(
            as.integer(config$labels$build$start_year),
            as.integer(config$labels$build$end_year)
          ),
          sep = ""
        ))
      )
    ),
    htmltools::div(
      class = "report-chart wide-report-chart",
      highcharter::highchartOutput(paste0(prefix, "_chart"), height = "520px")
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the selected NBD3-R values"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download selected NBD3-R data",
          class = "download-link"
        )
      )
    )
  )
}

register_final_panel <- function(prefix, geography_type_name, domain_name, default_series) {
  ids <- list(
    geographies = paste0(prefix, "_geographies"),
    sex = paste0(prefix, "_sex"),
    cause = paste0(prefix, "_cause"),
    age = paste0(prefix, "_age"),
    measure = paste0(prefix, "_measure"),
    years = paste0(prefix, "_years"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  availability <- unique(catalog$availability[
    model == "NBD3-R" &
      get("geography_type") %in% geography_type_name &
      get("domain") == domain_name
  ])
  age_catalog <- unique(availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]

  shiny::observeEvent(input[[ids$measure]], {
    selected_measure <- input[[ids$measure]] %||% "crude_rate"
    choices <- if (identical(selected_measure, "asr")) {
      age_catalog[age_id == "asr_all"]
    } else if (identical(selected_measure, "crude_rate")) {
      age_catalog[age_id != "asr_all" & age_id != "age_0"]
    } else {
      age_catalog[age_id != "asr_all"]
    }
    preferred <- if (identical(selected_measure, "asr")) "asr_all" else "age_all"
    shiny::updateSelectizeInput(
      session,
      ids$age,
      choices = named_choices(choices, "age_id", "age_label"),
      selected = safe_default(choices$age_id, preferred),
      server = TRUE
    )
  }, ignoreInit = FALSE)

  selected_data <- shiny::reactive({
    shiny::req(
      input[[ids$geographies]], input[[ids$sex]], input[[ids$cause]],
      input[[ids$age]], input[[ids$measure]], input[[ids$years]]
    )
    selected_geography_codes <- suppressWarnings(as.integer(unlist(
      input[[ids$geographies]], recursive = TRUE, use.names = FALSE
    )))
    selected_geography_codes <- unique(selected_geography_codes[!is.na(selected_geography_codes)])
    selected_sex_code <- suppressWarnings(as.integer(unlist(
      input[[ids$sex]], recursive = TRUE, use.names = FALSE
    )))[[1L]]
    selected_years <- suppressWarnings(as.integer(unlist(
      input[[ids$years]], recursive = TRUE, use.names = FALSE
    )))
    selected_years <- selected_years[!is.na(selected_years)]

    d <- collect_cause_rates(
      runtime$cause_rates_dataset,
      models = "NBD3-R",
      geography_type = geography_type_name,
      geography_codes = selected_geography_codes,
      sex_code = selected_sex_code,
      series_id = input[[ids$cause]],
      age_id = input[[ids$age]],
      year_range = selected_years,
      measure = input[[ids$measure]]
    )
    if (!nrow(d)) return(d)
    d <- attach_exact_cause_uncertainty(d)
    d[, line_name := geography]
    d[order(geography_code, year)]
  })

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 520))
    hc_nbd_lines(
      d,
      y = "estimate",
      series = "line_name",
      lower = "lower",
      upper = "upper",
      title = unique(d$series_label)[[1L]],
      subtitle = paste(unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = measure_label(input[[ids$measure]], config),
      config = config,
      decimals = metric_decimals(input[[ids$measure]]),
      height = 520,
      filename = paste0(prefix, "-nbd3-r-trend")
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No NBD3-R results are available for this selection."
      ))
    }
    interval_rows <- d[is.finite(lower) & is.finite(upper), .N]
    total_rows <- nrow(d)
    interval_message <- if (interval_rows > 0L) {
      paste0(
        format(interval_rows, big.mark = ","), " of ",
        format(total_rows, big.mark = ","),
        " plotted cells have 95% uncertainty intervals from ",
        uncertainty_draw_label, "."
      )
    } else {
      paste(
        "No full-grid interval was available for this cell. Confirm that the",
        "current uncertainty run completed all province and population-group",
        "full-draw files for Male, Female and Person."
      )
    }
    htmltools::div(
      class = if (interval_rows > 0L) "narrative-box success" else "narrative-box",
      htmltools::strong("NBD3-R implementation. "),
      interval_message,
      htmltools::br(),
      htmltools::span(
        class = "provisional-note",
        paste(
          "Intervals combine under-reporting, injury-envelope, injury",
          "survey-composition, HIV fitted-parameter, and redistribution uncertainty in one joint run."
        )
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    selected_measure <- input[[ids$measure]]
    digits <- metric_decimals(selected_measure)
    display <- d[, .(
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Cause = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      `Draws used` = n_draws
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, selected_measure, digits),
      Lower = format_table_value(Lower, selected_measure, digits),
      Upper = format_table_value(Upper, selected_measure, digits),
      `Draws used` = vapply(
        `Draws used`,
        function(value) if (is.na(value)) "—" else format(value, big.mark = ","),
        character(1)
      )
    )]
    nbd_kable_html(
      display,
      caption = "Selected NBD3-R estimates and 95% uncertainty intervals",
      align = c("l", "l", "r", "l", "l", "l", "r", "r", "r", "r")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-nbd3-r.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )

  invisible(TRUE)
}

# Section 4.3: broad cause group × age profile -------------------------------
#
# The panel below uses the same optimized point/uncertainty cache as the two
# general result explorers, but it requests point-estimate deaths only. Four
# fixed years are displayed in a 2 × 2 layout, with the age-specific deaths
# stacked into the four broad cause groups used in the published figure.

broad_age_profile_years <- c(1997L, 2005L, 2012L, 2019L)

broad_age_profile_ages <- data.table::data.table(
  age_id = c(
    "age_under1", "age_1_4", "age_5_9", "age_10_14", "age_15_19",
    "age_20_24", "age_25_29", "age_30_34", "age_35_39", "age_40_44",
    "age_45_49", "age_50_54", "age_55_59", "age_60_64", "age_65_69",
    "age_70_74", "age_75_79", "age_80_84", "age_85_plus"
  ),
  age_label = c(
    "0", "1–4", "5–9", "10–14", "15–19",
    "20–24", "25–29", "30–34", "35–39", "40–44",
    "45–49", "50–54", "55–59", "60–64", "65–69",
    "70–74", "75–79", "80–84", "85+"
  ),
  age_order = seq_len(19L)
)

# The colours are sampled from the reference figure supplied for this report.
# Stack order is bottom to top; legend order is the reverse, matching the
# reference figure.
broad_age_profile_causes <- data.table::data.table(
  series_id = c("za_168", "za_169", "za_170", "za_171"),
  series_label = c(
    paste(
      "Communicable diseases, perinatal conditions, maternal causes,",
      "and nutritional deficiencies"
    ),
    "HIV/AIDS and tuberculosis",
    "Non-communicable diseases",
    "Injuries"
  ),
  legend_label = c(
    paste(
      "Communicable diseases, perinatal conditions, maternal causes,",
      "and nutritional deficiencies"
    ),
    "HIV/AIDS and tuberculosis",
    "Non-communicable diseases",
    "Injuries"
  ),
  colour = c("#926DB2", "#AC0639", "#3B9FC2", "#72B43C"),
  stack_order = 1:4,
  legend_order = c(4L, 3L, 2L, 1L)
)

resolve_broad_age_profile_causes <- function() {
  causes <- data.table::copy(broad_age_profile_causes)
  available_ids <- unique(as.character(result_cause_catalog$series_id))
  if (all(causes$series_id %in% available_ids)) return(causes)

  # Stable label-based fallback for repositories whose report IDs were
  # regenerated while retaining the same cause hierarchy.
  fallback_patterns <- c(
    "communicable.*maternal.*perinatal|comm/mat/peri/nutr",
    "hiv/aids.*tuberculosis|hiv/aids.*tb",
    "non-communicable",
    "^injuries$"
  )
  for (i in seq_len(nrow(causes))) {
    if (causes$series_id[[i]] %in% available_ids) next
    hit <- result_cause_catalog[
      grepl(
        fallback_patterns[[i]],
        series_label,
        ignore.case = TRUE,
        perl = TRUE
      )
    ]
    if (nrow(hit)) causes$series_id[[i]] <- hit$series_id[[1L]]
  }
  causes
}

broad_age_profile_panel_chart <- function(
    data,
    panel_year,
    y_max,
    show_y_title = FALSE,
    show_x_title = FALSE,
    config = NULL,
    height = 355) {
  d <- data.table::as.data.table(data.table::copy(data))
  if (!nrow(d)) return(empty_nbd_chart(height = height))

  causes <- resolve_broad_age_profile_causes()[order(stack_order)]
  total_deaths <- sum(d$estimate, na.rm = TRUE)
  panel_title <- paste0(
    panel_year,
    " (N=",
    format(round(total_deaths), big.mark = ",", scientific = FALSE),
    ")"
  )

  complete_grid <- data.table::CJ(
    age_id = broad_age_profile_ages$age_id,
    series_id = causes$series_id,
    unique = TRUE
  )
  d <- merge(
    complete_grid,
    d[, .(age_id, series_id, estimate)],
    by = c("age_id", "series_id"),
    all.x = TRUE,
    sort = FALSE
  )
  d[!is.finite(estimate), estimate := 0]
  d <- merge(d, broad_age_profile_ages, by = "age_id", all.x = TRUE)
  d <- merge(d, causes, by = "series_id", all.x = TRUE)
  data.table::setorder(d, stack_order, age_order)

  hc <- highcharter::highchart() |>
    highcharter::hc_chart(
      type = "column",
      spacing = c(8, 8, 8, 8),
      marginTop = 42
    ) |>
    highcharter::hc_title(
      text = panel_title,
      align = "left",
      x = 4,
      y = 15,
      style = list(
        color = "#30343b",
        fontSize = "13px",
        fontWeight = "600"
      )
    ) |>
    highcharter::hc_xAxis(
      categories = broad_age_profile_ages$age_label,
      title = list(text = if (show_x_title) "Age group (years)" else NULL),
      labels = list(
        rotation = -45,
        style = list(fontSize = "10px", color = "#4d535a")
      ),
      tickLength = 4,
      lineWidth = 1,
      lineColor = "#4d535a",
      tickColor = "#4d535a"
    ) |>
    highcharter::hc_yAxis(
      min = 0,
      max = y_max,
      tickAmount = 6,
      reversedStacks = FALSE,
      title = list(text = if (show_y_title) "Number of deaths" else NULL),
      gridLineColor = "#d8d8d8",
      gridLineWidth = 1,
      labels = list(style = list(fontSize = "10px", color = "#4d535a"))
    ) |>
    highcharter::hc_tooltip(
      shared = TRUE,
      useHTML = TRUE,
      headerFormat = "<b>{point.key}</b><br/>",
      pointFormat = paste0(
        '<span style="color:{series.color}">●</span> ',
        '{series.name}: <b>{point.y:,.0f}</b><br/>'
      ),
      footerFormat = "<b>Total: {point.stackTotal:,.0f}</b>"
    ) |>
    highcharter::hc_plotOptions(
      column = list(
        stacking = "normal",
        borderColor = "#4b4b4b",
        borderWidth = 0.6,
        pointPadding = 0.02,
        groupPadding = 0.05
      ),
      series = list(
        animation = nbd_hc_animation(duration = 520L),
        animationLimit = 10000L,
        states = list(inactive = list(opacity = 0.22))
      )
    ) |>
    highcharter::hc_legend(enabled = FALSE)

  for (i in seq_len(nrow(causes))) {
    cause <- causes[i]
    values <- d[series_id == cause$series_id][order(age_order)]$estimate
    hc <- highcharter::hc_add_series(
      hc,
      type = "column",
      name = cause$legend_label,
      data = as.numeric(values),
      color = cause$colour,
      borderColor = "#4b4b4b",
      borderWidth = 0.6,
      showInLegend = FALSE,
      animation = nbd_hc_animation(
        duration = 520L,
        defer = 70L * (i - 1L)
      )
    )
  }

  finish_nbd_hc(
    hc,
    config = config,
    height = height,
    filename = paste0("broad-cause-age-", panel_year)
  )
}

broad_age_profile_legend <- function() {
  causes <- resolve_broad_age_profile_causes()[order(legend_order)]
  htmltools::div(
    class = "broad-age-profile-legend",
    style = paste(
      "display:flex;flex-wrap:wrap;justify-content:center;",
      "gap:0.75rem 1.35rem;margin:0.2rem 0 1rem 0;",
      "color:#29465b;font-size:0.92rem;font-weight:500;"
    ),
    lapply(seq_len(nrow(causes)), function(i) {
      htmltools::span(
        class = "broad-age-profile-legend-item",
        style = "display:inline-flex;align-items:center;gap:0.42rem;",
        htmltools::span(
          class = "broad-age-profile-swatch",
          style = paste0(
            "display:inline-block;width:14px;height:14px;",
            "border:1px solid #4b4b4b;background:",
            causes$colour[[i]], ";"
          )
        ),
        causes$legend_label[[i]]
      )
    })
  )
}

broad_age_profile_panel_ui <- function(prefix = "result_broad_age") {
  ids <- paste0(prefix, c("_scope", "_geography", "_sex"))
  sexes <- unique(result_availability[, .(sex_code, sex)])[order(sex_code)]
  province_choices <- geography_choices_for_scope("province")

  shiny::tagList(
    htmltools::div(
      class = "control-panel research-control-panel",
      shiny::fluidRow(
        shiny::column(3, shiny::selectInput(
          ids[[1L]], "Population set",
          choices = c(
            "South Africa or province" = "province",
            "South Africa or population group" = "population_group"
          ),
          selected = "province"
        )),
        shiny::column(6, shiny::selectizeInput(
          ids[[2L]], "Geography or population group",
          choices = province_choices,
          selected = make_geography_key("national", 10L),
          multiple = FALSE
        )),
        shiny::column(3, shiny::selectizeInput(
          ids[[3L]], "Sex",
          choices = named_choices(sexes, "sex_code", "sex"),
          selected = safe_default(sexes$sex_code, 3L),
          multiple = FALSE
        ))
      )
    ),
    htmltools::div(
      class = "broad-age-profile-grid",
      shiny::fluidRow(
        shiny::column(
          6,
          htmltools::div(
            class = "report-chart broad-age-profile-panel",
            highcharter::highchartOutput(
              paste0(prefix, "_chart_1997"),
              height = "355px"
            )
          )
        ),
        shiny::column(
          6,
          htmltools::div(
            class = "report-chart broad-age-profile-panel",
            highcharter::highchartOutput(
              paste0(prefix, "_chart_2005"),
              height = "355px"
            )
          )
        )
      ),
      shiny::fluidRow(
        shiny::column(
          6,
          htmltools::div(
            class = "report-chart broad-age-profile-panel",
            highcharter::highchartOutput(
              paste0(prefix, "_chart_2012"),
              height = "355px"
            )
          )
        ),
        shiny::column(
          6,
          htmltools::div(
            class = "report-chart broad-age-profile-panel",
            highcharter::highchartOutput(
              paste0(prefix, "_chart_2019"),
              height = "355px"
            )
          )
        )
      )
    ),
    broad_age_profile_legend(),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary(
        "View and download the broad-cause age profile"
      ),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download broad-cause age profile",
          class = "download-link"
        )
      )
    )
  )
}

register_broad_age_profile_panel <- function(prefix = "result_broad_age") {
  ids <- list(
    scope = paste0(prefix, "_scope"),
    geography = paste0(prefix, "_geography"),
    sex = paste0(prefix, "_sex"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )
  chart_ids <- stats::setNames(
    paste0(prefix, "_chart_", broad_age_profile_years),
    broad_age_profile_years
  )

  shiny::observeEvent(input[[ids$scope]], {
    scope <- input[[ids$scope]] %||% "province"
    choices <- geography_choices_for_scope(scope)
    selected <- make_geography_key("national", 10L)
    if (!selected %in% unname(choices)) selected <- unname(choices)[[1L]]
    shiny::updateSelectizeInput(
      session,
      ids$geography,
      choices = choices,
      selected = selected,
      server = TRUE
    )
  }, ignoreInit = TRUE)

  selection_key <- shiny::reactive({
    shiny::req(input[[ids$geography]], input[[ids$sex]])
    list(
      panel = prefix,
      geography = as.character(input[[ids$geography]]),
      sex = as.integer(input[[ids$sex]])
    )
  })

  selected_data <- nbd_cached_reactive(
    selection_key,
    function(key) {
      causes <- resolve_broad_age_profile_causes()
      pieces <- lapply(
        seq_len(nrow(broad_age_profile_ages)),
        function(i) {
          age <- broad_age_profile_ages[i]
          d <- collect_result_slice(
            geography_keys = key$geography,
            sex_code = key$sex,
            series_ids = causes$series_id,
            age_id = age$age_id,
            year_range = range(broad_age_profile_years),
            measure = "deaths"
          )
          if (!nrow(d)) return(data.table::data.table())
          d <- d[year %in% broad_age_profile_years]
          d[, `:=`(
            age_id = age$age_id,
            age_label = age$age_label,
            age_order = age$age_order
          )]
          d
        }
      )
      d <- data.table::rbindlist(pieces, use.names = TRUE, fill = TRUE)
      if (!nrow(d)) return(d)
      d <- merge(
        d,
        causes[, .(
          series_id,
          broad_label = series_label,
          colour,
          stack_order,
          legend_order
        )],
        by = "series_id",
        all.x = TRUE,
        sort = FALSE
      )
      data.table::setorder(d, year, age_order, stack_order)
      d
    },
    millis = 250L
  )

  y_max <- shiny::reactive({
    d <- selected_data()
    if (!nrow(d)) return(NULL)
    totals <- d[, .(age_total = sum(estimate, na.rm = TRUE)),
                by = .(year, age_id)]
    maximum <- max(totals$age_total, na.rm = TRUE)
    if (!is.finite(maximum) || maximum <= 0) return(NULL)
    magnitude <- 10^floor(log10(maximum))
    step <- if (maximum / magnitude < 2) {
      magnitude / 5
    } else if (maximum / magnitude < 5) {
      magnitude / 2
    } else {
      magnitude
    }
    ceiling(maximum / step) * step
  })

  for (panel_year in broad_age_profile_years) {
    local({
      year_value <- panel_year
      output[[chart_ids[[as.character(year_value)]]]] <-
        highcharter::renderHighchart({
          d <- data.table::copy(selected_data())
          ymax <- y_max()
          if (!nrow(d) || is.null(ymax)) {
            return(empty_nbd_chart(height = 355))
          }
          broad_age_profile_panel_chart(
            d[year == year_value],
            panel_year = year_value,
            y_max = ymax,
            show_y_title = year_value %in% c(1997L, 2012L),
            show_x_title = year_value %in% c(2012L, 2019L),
            config = config,
            height = 355
          )
        })
    })
  }

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No broad-cause age profile is available for this selection."
      ))
    }
    htmltools::div(
      class = "narrative-box",
      htmltools::strong("Broad cause composition by age. "),
      paste0(
        "The four panels show deaths in 1997, 2005, 2012, and 2019 for ",
        unique(d$geography)[[1L]], " — ", unique(d$sex)[[1L]],
        ". The four stacked groups sum to the displayed all-cause total in ",
        "each age group."
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    display <- d[, .(
      Geography = geography,
      Sex = sex,
      Year = year,
      `Age group` = age_label,
      `Broad cause group` = broad_label,
      Deaths = round(estimate)
    )]
    nbd_kable_html(
      display,
      caption = paste(
        "Deaths by broad cause group and age group for the four selected years"
      ),
      align = c("l", "l", "r", "l", "l", "r")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-broad-cause-age-profile.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )

  invisible(TRUE)
}


# Injury source table used in the methods section.
injury_source_table <- data.table::data.table()
source_fractions <- data.table::as.data.table(data.table::copy(
  injury_inputs$source_fractions
))
if (nrow(source_fractions) && all(c(
  "survey", "year", "nbdcode", "observed_count"
) %in% names(source_fractions))) {
  labels <- data.table::fread(file.path(root, "data", "lookups", "analysis_codes.csv"))
  injury_source_table <- source_fractions[, .(
    observed_deaths = sum(observed_count, na.rm = TRUE)
  ), by = .(survey, year, nbdcode)]
  injury_source_table[, percent := safe_ratio(
    observed_deaths, sum(observed_deaths), 100
  ), by = .(survey, year)]
  injury_source_table <- merge(
    injury_source_table,
    labels[, .(nbdcode = as.integer(analysis_code), Cause = analysis_label)],
    by = "nbdcode",
    all.x = TRUE,
    sort = FALSE
  )
  injury_source_table[is.na(Cause), Cause := paste("Cause", nbdcode)]
  injury_source_table[, display := paste0(
    "<span class=\"injury-table-count\">",
    format(round(observed_deaths), big.mark = ",", scientific = FALSE),
    "</span><span class=\"injury-table-share\">",
    sprintf("%.2f%%", percent),
    "</span>"
  )]
  injury_source_table <- data.table::dcast(
    injury_source_table,
    nbdcode + Cause ~ survey,
    value.var = "display",
    fill = "—"
  )
  data.table::setorder(injury_source_table, nbdcode)
  data.table::setnames(
    injury_source_table,
    c("nbdcode", "Cause"),
    c("Code", "Cause of injury")
  )
  survey_columns <- intersect(
    c("NIMS 2000", "IMS 2009", "FAMHIS 2017"),
    names(injury_source_table)
  )
  data.table::setcolorder(
    injury_source_table,
    c("Code", "Cause of injury", survey_columns)
  )
}

injury_cause_labels <- data.table::fread(
  file.path(root, "data", "lookups", "analysis_codes.csv")
)[, .(
  nbdcode = as.integer(analysis_code),
  Cause = as.character(analysis_label)
)]

injury_broad_fit_table <- data.table::data.table()
if (nrow(injury_broad_comparison)) {
  injury_broad_fit_table <- injury_broad_comparison[, .(
    Survey = survey,
    Year = year,
    `Broad injury group` = broad_group,
    `Survey input (%)` = 100 * survey_fraction,
    `Final smoothed estimate (%)` = 100 * model_fraction,
    `Difference (percentage points)` = difference_percentage_points
  )]
  data.table::setorderv(
    injury_broad_fit_table,
    c("Year", "Broad injury group")
  )
}

injury_flagged_fit_table <- data.table::data.table()
if (nrow(injury_detailed_comparison)) {
  injury_flagged_fit_table <- merge(
    injury_detailed_comparison[nbdcode %in% c(132L, 136L, 138L)],
    injury_cause_labels,
    by = "nbdcode",
    all.x = TRUE,
    sort = FALSE
  )[, .(
    Survey = survey,
    Year = year,
    Code = nbdcode,
    Cause,
    `Survey input (%)` = 100 * survey_fraction,
    `Final smoothed estimate (%)` = 100 * model_fraction,
    `Difference (percentage points)` = difference_percentage_points
  )]
  data.table::setorder(injury_flagged_fit_table, Year, Code)
}

# -----------------------------------------------------------------------------
# Research-report additions: completeness diagnostics and two general result
# explorers. These are report-layer helpers only; they do not alter estimates.
# -----------------------------------------------------------------------------
report_fallback_colours <- c(
  "#0071BC", "#1B998B", "#D48A16", "#B23A48", "#7A5195",
  "#5B8E7D", "#2F4858", "#E76F51", "#6C757D", "#8A5A44",
  "#3A86FF", "#8338EC"
)
series_colour <- function(series, config, index = 1L) {
  colours <- unlist(config$labels$models$colours, use.names = TRUE)
  if (series %in% names(colours)) return(unname(colours[[series]]))
  report_fallback_colours[[1L + ((index - 1L) %% length(report_fallback_colours))]]
}
read_optional_report_table <- function(path) {
  if (!length(path) || is.na(path) || !nzchar(path) || !file.exists(path)) {
    return(data.table::data.table())
  }
  tryCatch({
    extension <- tolower(tools::file_ext(path))
    if (identical(extension, "parquet")) {
      data.table::as.data.table(arrow::read_parquet(path))
    } else {
      data.table::fread(path)
    }
  }, error = function(error) {
    warning("Could not read optional report input: ", path, ". ", conditionMessage(error))
    data.table::data.table()
  })
}

analysis_config <- yaml::read_yaml(file.path(root, "config", "config.yml"))
uncertainty_config <- yaml::read_yaml(
  file.path(root, "config", "uncertainty_joint.yml")
)

find_uncertainty_report_root <- function() {
  configured_name <- uncertainty_config$run$output_name %||% ""
  configured_path <- file.path(root, "output", "uncertainty", configured_name)
  if (nzchar(configured_name) && dir.exists(configured_path)) {
    return(configured_path)
  }
  base <- file.path(root, "output", "uncertainty")
  if (!dir.exists(base)) return(NA_character_)
  candidates <- list.dirs(base, recursive = FALSE, full.names = TRUE)
  candidates <- candidates[file.exists(file.path(
    candidates, "completeness_weighted_cells.parquet"
  ))]
  if (!length(candidates)) return(NA_character_)
  candidates[[which.max(file.info(candidates)$mtime)]]
}

uncertainty_report_root <- find_uncertainty_report_root()
report_injury_codes <- c(
  124L, 125L, 127L, 128L, 129L, 130L, 131L, 132L,
  135L, 136L, 137L, 138L, 139L, 140L, 141L
)
report_province_labels <- c(
  `1` = "Western Cape", `2` = "Eastern Cape", `3` = "Northern Cape",
  `4` = "Free State", `5` = "KwaZulu-Natal", `6` = "North West",
  `7` = "Gauteng", `8` = "Mpumalanga", `9` = "Limpopo"
)

# Prefer the exact annual cells written by the uncertainty engine. If they are
# not yet available, reconstruct the same death-weighted province-year S2 view
# from the deterministic Stage 04 checkpoints.
completeness_time_data <- read_optional_report_table(report_method_asset(
  config,
  "completeness_weighted_cells.parquet",
  fallback = if (!is.na(uncertainty_report_root)) file.path(
    uncertainty_report_root, "completeness_weighted_cells.parquet"
  ) else NA_character_
))

if (!nrow(completeness_time_data)) {
  scalar_path <- file.path(
    root, "data", "derived", "04_completeness_scalars.parquet"
  )
  stage04_candidates <- c(
    file.path(
      root, "data", "derived",
      "04_investigation_subpopulation_pre_injury_envelope.parquet"
    ),
    file.path(
      root, "data", "derived", "04_investigation_subpopulation.parquet"
    )
  )
  stage04_path <- stage04_candidates[file.exists(stage04_candidates)][1L]
  completeness_scalars <- read_optional_report_table(scalar_path)
  completeness_stage04 <- read_optional_report_table(stage04_path)
  required_scalar <- c("Death_Prov", "Sex", "DeathYear", "age5", "S2")
  required_stage04 <- c(
    "Death_Prov", "Sex", "DeathYear", "age5", "Popgroup",
    "nbdcode", "Deaths"
  )
  if (all(required_scalar %in% names(completeness_scalars)) &&
      all(required_stage04 %in% names(completeness_stage04))) {
    freeze_year <- as.integer(
      analysis_config$settings$completeness_freeze_year %||% 2010L
    )
    african_natural <- completeness_stage04[
      Popgroup == 1L & !nbdcode %in% report_injury_codes,
      .(adjusted_african_natural = sum(Deaths, na.rm = TRUE)),
      by = .(Death_Prov, Sex, DeathYear, age5)
    ]
    candidate <- merge(
      completeness_scalars[, .(
        Death_Prov = as.integer(Death_Prov),
        Sex = as.integer(Sex),
        DeathYear = as.integer(DeathYear),
        age5 = as.integer(age5),
        S2 = as.numeric(S2)
      )],
      african_natural,
      by = c("Death_Prov", "Sex", "DeathYear", "age5"),
      all.x = TRUE,
      sort = FALSE
    )
    candidate <- candidate[
      DeathYear <= freeze_year & age5 != 1L &
        is.finite(S2) & S2 > 0 &
        is.finite(adjusted_african_natural) &
        adjusted_african_natural > 0
    ]
    candidate[, pre_adjustment_african_natural :=
                adjusted_african_natural / S2]
    completeness_time_data <- candidate[
      is.finite(pre_adjustment_african_natural) &
        pre_adjustment_african_natural > 0,
      .(
        contributing_sex_age_cells = .N,
        contributing_sexes = data.table::uniqueN(Sex),
        contributing_age_groups = data.table::uniqueN(age5),
        adjusted_african_natural = sum(adjusted_african_natural),
        pre_adjustment_african_natural = sum(
          pre_adjustment_african_natural
        )
      ),
      by = .(Death_Prov, DeathYear)
    ]
    completeness_time_data[, S2 :=
      adjusted_african_natural / pre_adjustment_african_natural]
    completeness_time_data[, log_s2 := log(S2)]
  }
}

if (nrow(completeness_time_data)) {
  completeness_time_data[, `:=`(
    Death_Prov = as.integer(Death_Prov),
    DeathYear = as.integer(DeathYear),
    S2 = as.numeric(S2)
  )]
  completeness_time_data[, Province := unname(
    report_province_labels[as.character(Death_Prov)]
  )]
  completeness_time_data[is.na(Province), Province := paste(
    "Province", Death_Prov
  )]
  completeness_time_data[, implied_completeness := 100 / S2]
  data.table::setorder(completeness_time_data, Death_Prov, DeathYear)
}

completeness_province_data <- read_optional_report_table(report_method_asset(
  config,
  "completeness_by_province.csv",
  fallback = if (!is.na(uncertainty_report_root)) file.path(
    uncertainty_report_root, "completeness_by_province.csv"
  ) else NA_character_
))

if (!nrow(completeness_province_data) && nrow(completeness_time_data)) {
  completeness_province_data <- completeness_time_data[, {
    weights <- as.numeric(pre_adjustment_african_natural)
    values <- as.numeric(log(S2))
    weighted_mean_log <- sum(weights * values) / sum(weights)
    list(
      eligible_years = data.table::uniqueN(DeathYear),
      pre_adjustment_deaths = sum(pre_adjustment_african_natural),
      adjusted_deaths = sum(adjusted_african_natural),
      aggregate_s2 = sum(adjusted_african_natural) /
        sum(pre_adjustment_african_natural),
      log_sd = sqrt(sum(weights * (values - weighted_mean_log)^2) /
        sum(weights)),
      minimum_annual_s2 = min(S2),
      median_annual_s2 = stats::median(S2),
      maximum_annual_s2 = max(S2)
    )
  }, by = Death_Prov]
}

if (nrow(completeness_province_data)) {
  completeness_province_data[, Death_Prov := as.integer(Death_Prov)]
  completeness_province_data[, Province := unname(
    report_province_labels[as.character(Death_Prov)]
  )]
  completeness_province_data[is.na(Province), Province := paste(
    "Province", Death_Prov
  )]
  for (column in c(
    "aggregate_s2", "minimum_annual_s2", "median_annual_s2",
    "maximum_annual_s2", "log_sd"
  )) {
    if (!column %in% names(completeness_province_data)) {
      completeness_province_data[, (column) := NA_real_]
    }
  }
  completeness_province_data[, `:=`(
    implied_completeness = 100 / aggregate_s2,
    minimum_annual_completeness = 100 / maximum_annual_s2,
    median_annual_completeness = 100 / median_annual_s2,
    maximum_annual_completeness = 100 / minimum_annual_s2
  )]
  data.table::setorder(completeness_province_data, Death_Prov)
}

injury_anchor_data <- read_optional_report_table(report_method_asset(
  config,
  "04_injury_envelope_anchors.parquet",
  fallback = file.path(
    root, "data", "derived", "04_injury_envelope_anchors.parquet"
  )
))
if (nrow(injury_anchor_data)) {
  anchor_columns <- intersect(c(
    "survey_year", "survey", "survey_level_total", "routine_level_total",
    "published_survey_level_total", "published_survey_level_lower",
    "published_survey_level_upper", "derived_survey_level_source"
  ), names(injury_anchor_data))
  injury_anchor_data <- unique(injury_anchor_data[, ..anchor_columns])
  data.table::setorder(injury_anchor_data, survey_year)
}

# Final-result catalog used by the two general result explorers.
result_availability <- unique(catalog$availability[
  model == "NBD3-R" &
    geography_type %in% c("national", "province", "population_group")
])
result_cause_catalog <- unique(result_availability[, .(
  series_id, series_label, hierarchy, domain, cause_type, series_sort_order
)])[order(series_sort_order)]

report_hierarchy_labels <- c(
  broad = "All causes and broad causes",
  reporting = "Reporting aggregates",
  group = "Disease categories",
  detailed = "Detailed causes",
  custom = "Custom aggregates"
)
result_hierarchies <- intersect(
  names(report_hierarchy_labels), unique(result_cause_catalog$hierarchy)
)
result_hierarchy_choices <- c(
  "All hierarchy levels" = "all",
  stats::setNames(result_hierarchies, report_hierarchy_labels[result_hierarchies])
)

cause_choices_for_hierarchy <- function(hierarchy = "all", grouped = FALSE) {
  hierarchy_value <- as.character(hierarchy)[[1L]]
  causes <- data.table::copy(result_cause_catalog)
  if (!identical(hierarchy_value, "all")) {
    causes <- result_cause_catalog[get("hierarchy") == hierarchy_value]
  }
  if (!nrow(causes)) return(character())
  if (!grouped || !identical(hierarchy_value, "all")) {
    return(named_choices(causes, "series_id", "series_label"))
  }
  groups <- split(causes, causes$hierarchy)
  groups <- groups[names(groups) %in% result_hierarchies]
  out <- lapply(groups, named_choices, value = "series_id", label = "series_label")
  names(out) <- unname(report_hierarchy_labels[names(out)])
  out
}

default_cause_ids <- function(hierarchy = "all", multiple = FALSE) {
  available <- result_cause_catalog$series_id
  preferred <- if (multiple) {
    if (identical(hierarchy, "all")) {
      c("za_171", "za_2", "za_1")
    } else if (identical(hierarchy, "broad")) {
      c("za_172", "za_168", "za_169", "za_170", "za_171")
    } else if (identical(hierarchy, "group")) {
      c("za_150", "za_158", "za_166", "za_167")
    } else if (identical(hierarchy, "detailed")) {
      c("za_2", "za_1", "za_126", "za_139")
    } else {
      character()
    }
  } else {
    if (identical(hierarchy, "detailed")) "za_2" else "za_172"
  }
  selected <- preferred[preferred %in% available]
  if (length(selected)) return(selected)
  hierarchy_value <- as.character(hierarchy)[[1L]]
  causes <- if (identical(hierarchy_value, "all")) {
    result_cause_catalog
  } else {
    result_cause_catalog[get("hierarchy") == hierarchy_value]
  }
  if (!nrow(causes)) return(character())
  if (multiple) head(causes$series_id, 5L) else causes$series_id[[1L]]
}

make_geography_key <- function(type, code) paste(type, as.integer(code), sep = "::")
parse_geography_key <- function(key) {
  pieces <- strsplit(as.character(key), "::", fixed = TRUE)[[1L]]
  if (length(pieces) != 2L) return(list(type = NA_character_, code = NA_integer_))
  list(type = pieces[[1L]], code = suppressWarnings(as.integer(pieces[[2L]])))
}

geography_choices_for_scope <- function(scope = "province") {
  types <- if (identical(scope, "population_group")) {
    c("national", "population_group")
  } else {
    c("national", "province")
  }
  geographies <- unique(result_availability[
    geography_type %in% types,
    .(geography_type, geography_code, geography)
  ])
  geographies[, key := make_geography_key(geography_type, geography_code)]
  geographies[, label := geography]
  geographies[, type_order := match(
    geography_type, c("national", "province", "population_group")
  )]
  data.table::setorder(geographies, type_order, geography_code)
  stats::setNames(geographies$key, geographies$label)
}

attach_exact_cause_uncertainty <- function(data) {
  d <- data.table::as.data.table(data.table::copy(data))
  if (!nrow(d)) return(d)
  key <- c(
    "geography_type", "geography_code", "sex_code", "year",
    "age_id", "series_id", "measure"
  )
  d[, `:=`(
    lower = NA_real_, upper = NA_real_, n_draws = NA_integer_,
    uncertainty_source = NA_character_
  )]

  dynamic <- tryCatch(
    collect_full_cause_uncertainty(runtime, d),
    error = function(error) {
      warning(
        "Full-grid uncertainty could not be calculated: ",
        conditionMessage(error),
        call. = FALSE
      )
      data.table::data.table()
    }
  )
  if (nrow(dynamic)) {
    dynamic <- dynamic[, c(key, "lower", "upper", "n_draws", "source"),
                       with = FALSE]
    data.table::setnames(dynamic, "source", "uncertainty_source")
    d[, c("lower", "upper", "n_draws", "uncertainty_source") := NULL]
    return(merge(d, dynamic, by = key, all.x = TRUE, sort = FALSE))
  }

  # Backward-compatible fallback for an older compact uncertainty run.
  if (!nrow(cause_uncertainty)) return(d)
  u <- cause_uncertainty[
    geography_type %in% unique(d$geography_type) &
      geography_code %in% unique(d$geography_code) &
      sex_code %in% unique(d$sex_code) &
      year %in% unique(d$year) &
      age_id %in% unique(d$age_id) &
      series_id %in% unique(d$series_id) &
      measure %in% unique(d$measure),
    c(key, "lower", "upper", "n_draws", "source"),
    with = FALSE
  ]
  if (!nrow(u)) return(d)
  data.table::setnames(u, "source", "uncertainty_source")
  d[, c("lower", "upper", "n_draws", "uncertainty_source") := NULL]
  merge(d, u, by = key, all.x = TRUE, sort = FALSE)
}


collect_result_slice <- function(
    geography_keys,
    sex_code,
    series_ids,
    age_id,
    year_range,
    measure) {
  geography_keys <- unique(as.character(geography_keys))
  series_ids <- unique(as.character(series_ids))

  # Production fast path: deterministic values and uncertainty limits are read
  # together from one small sex/age partition. No raw uncertainty draw is
  # scanned during the Shiny session.
  fast <- collect_fast_ui_results(
    runtime = runtime,
    geography_keys = geography_keys,
    sex_code = sex_code,
    series_ids = series_ids,
    age_id = age_id,
    year_range = year_range,
    measure = measure
  )
  if (nrow(fast)) return(fast)

  pieces <- lapply(geography_keys, function(geography_key) {
    geography <- parse_geography_key(geography_key)
    if (is.na(geography$type) || is.na(geography$code)) {
      return(data.table::data.table())
    }
    data.table::rbindlist(lapply(series_ids, function(series_id) {
      collect_cause_rates(
        runtime$cause_rates_dataset,
        models = "NBD3-R",
        geography_type = geography$type,
        geography_codes = geography$code,
        sex_code = sex_code,
        series_id = series_id,
        age_id = age_id,
        year_range = year_range,
        measure = measure
      )
    }), use.names = TRUE, fill = TRUE)
  })
  d <- data.table::rbindlist(pieces, use.names = TRUE, fill = TRUE)
  attach_exact_cause_uncertainty(d)
}

update_result_age_choices <- function(input_id, selected_measure) {
  ages <- unique(result_availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  choices <- if (identical(selected_measure, "asr")) {
    ages[age_id == "asr_all"]
  } else if (identical(selected_measure, "crude_rate")) {
    # Neonatal deaths are available, but the current population input has no
    # separate neonatal denominator. The combined under-one rate is supported.
    ages[age_id != "asr_all" & age_id != "age_0"]
  } else {
    ages[age_id != "asr_all"]
  }
  preferred <- if (identical(selected_measure, "asr")) "asr_all" else "age_all"
  shiny::updateSelectizeInput(
    session,
    input_id,
    choices = named_choices(choices, "age_id", "age_label"),
    selected = safe_default(choices$age_id, preferred),
    server = TRUE
  )
}

result_geography_panel_ui <- function(prefix = "result_geography") {
  ids <- paste0(prefix, c(
    "_scope", "_geographies", "_hierarchy", "_cause", "_sex",
    "_age", "_measure", "_years"
  ))
  ages <- unique(result_availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  sexes <- unique(result_availability[, .(sex_code, sex)])[order(sex_code)]
  shiny::tagList(
    htmltools::div(
      class = "control-panel research-control-panel",
      shiny::fluidRow(
        shiny::column(3, shiny::selectInput(
          ids[[1L]], "Comparison set",
          choices = c(
            "South Africa and provinces" = "province",
            "South Africa and population groups" = "population_group"
          ),
          selected = "province"
        )),
        shiny::column(5, shiny::selectizeInput(
          ids[[2L]], "Geographies",
          choices = geography_choices_for_scope("province"),
          selected = unname(geography_choices_for_scope("province")),
          multiple = TRUE
        )),
        shiny::column(4, shiny::selectInput(
          ids[[3L]], "Cause hierarchy",
          choices = result_hierarchy_choices,
          selected = "broad"
        ))
      ),
      shiny::fluidRow(
        shiny::column(4, shiny::selectizeInput(
          ids[[4L]], "Cause or aggregate",
          choices = cause_choices_for_hierarchy("broad"),
          selected = default_cause_ids("broad", multiple = FALSE),
          multiple = FALSE
        )),
        shiny::column(2, shiny::selectizeInput(
          ids[[5L]], "Sex",
          choices = named_choices(sexes, "sex_code", "sex"),
          selected = safe_default(sexes$sex_code, 3L),
          multiple = FALSE
        )),
        shiny::column(3, shiny::selectizeInput(
          ids[[6L]], "Age group",
          choices = named_choices(ages[age_id != "asr_all"], "age_id", "age_label"),
          selected = safe_default(ages$age_id, "age_all"),
          multiple = FALSE
        )),
        shiny::column(3, shiny::selectInput(
          ids[[7L]], "Measure",
          choices = measure_choices,
          selected = "crude_rate"
        ))
      ),
      shiny::sliderInput(
        ids[[8L]], "Years",
        min = as.integer(config$labels$build$start_year),
        max = as.integer(config$labels$build$end_year),
        value = c(
          as.integer(config$labels$build$start_year),
          as.integer(config$labels$build$end_year)
        ),
        sep = ""
      )
    ),
    htmltools::div(
      class = "report-chart wide-report-chart",
      highcharter::highchartOutput(paste0(prefix, "_chart"), height = "540px")
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the selected geography comparison"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download selected geography comparison",
          class = "download-link"
        )
      )
    )
  )
}

register_result_geography_panel <- function(prefix = "result_geography") {
  ids <- list(
    scope = paste0(prefix, "_scope"),
    geographies = paste0(prefix, "_geographies"),
    hierarchy = paste0(prefix, "_hierarchy"),
    cause = paste0(prefix, "_cause"),
    sex = paste0(prefix, "_sex"),
    age = paste0(prefix, "_age"),
    measure = paste0(prefix, "_measure"),
    years = paste0(prefix, "_years"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  shiny::observeEvent(input[[ids$scope]], {
    choices <- geography_choices_for_scope(input[[ids$scope]] %||% "province")
    shiny::updateSelectizeInput(
      session, ids$geographies,
      choices = choices,
      selected = unname(choices),
      server = TRUE
    )
  }, ignoreInit = TRUE)

  shiny::observeEvent(input[[ids$hierarchy]], {
    hierarchy <- input[[ids$hierarchy]] %||% "broad"
    shiny::updateSelectizeInput(
      session, ids$cause,
      choices = cause_choices_for_hierarchy(hierarchy, grouped = identical(hierarchy, "all")),
      selected = default_cause_ids(hierarchy, multiple = FALSE),
      server = TRUE
    )
  }, ignoreInit = TRUE)

  shiny::observeEvent(input[[ids$measure]], {
    update_result_age_choices(ids$age, input[[ids$measure]] %||% "crude_rate")
  }, ignoreInit = FALSE)

  selection_key <- shiny::reactive({
    shiny::req(
      input[[ids$geographies]], input[[ids$cause]], input[[ids$sex]],
      input[[ids$age]], input[[ids$measure]], input[[ids$years]]
    )
    list(
      panel = prefix,
      geographies = as.character(input[[ids$geographies]]),
      cause = as.character(input[[ids$cause]]),
      sex = as.integer(input[[ids$sex]]),
      age = as.character(input[[ids$age]]),
      measure = as.character(input[[ids$measure]]),
      years = as.integer(input[[ids$years]])
    )
  })

  selected_data <- nbd_cached_reactive(
    selection_key,
    function(key) {
      d <- collect_result_slice(
        geography_keys = key$geographies,
        sex_code = key$sex,
        series_ids = key$cause,
        age_id = key$age,
        year_range = key$years,
        measure = key$measure
      )
      if (nrow(d)) {
        d[, line_name := geography]
        data.table::setorder(d, geography_type, geography_code, year)
      }
      d
    },
    millis = 250L
  )

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 540))
    hc_nbd_lines(
      d,
      y = "estimate",
      series = "line_name",
      lower = "lower",
      upper = "upper",
      title = unique(d$series_label)[[1L]],
      subtitle = paste(unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = measure_label(input[[ids$measure]], config),
      config = config,
      decimals = metric_decimals(input[[ids$measure]]),
      height = 540,
      filename = paste0(prefix, "-geography-comparison")
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No NBD3-R result is available for this selection."
      ))
    }
    interval_rows <- d[is.finite(lower) & is.finite(upper), .N]
    htmltools::div(
      class = if (interval_rows) "narrative-box success" else "narrative-box",
      htmltools::strong("Geography comparison. "),
      paste0(
        data.table::uniqueN(d$line_name), " series are displayed; ",
        interval_rows, " of ", nrow(d),
        " plotted cells have exact 95% intervals from ", uncertainty_draw_label, "."
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    measure <- input[[ids$measure]]
    digits <- metric_decimals(measure)
    display <- d[, .(
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Cause = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      `Draws used` = n_draws
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, measure, digits),
      Lower = format_table_value(Lower, measure, digits),
      Upper = format_table_value(Upper, measure, digits),
      `Draws used` = ifelse(is.na(`Draws used`), "—", format(`Draws used`, big.mark = ","))
    )]
    nbd_kable_html(
      display,
      caption = "Selected NBD3-R values by geography or population group",
      align = c("l", "l", "r", "l", "l", "l", "r", "r", "r", "r")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-geography-comparison.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )
  invisible(TRUE)
}

result_cause_panel_ui <- function(prefix = "result_cause") {
  ids <- paste0(prefix, c(
    "_scope", "_geography", "_hierarchy", "_causes", "_sex",
    "_age", "_measure", "_years"
  ))
  ages <- unique(result_availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  sexes <- unique(result_availability[, .(sex_code, sex)])[order(sex_code)]
  province_choices <- geography_choices_for_scope("province")
  shiny::tagList(
    htmltools::div(
      class = "control-panel research-control-panel",
      shiny::fluidRow(
        shiny::column(3, shiny::selectInput(
          ids[[1L]], "Population set",
          choices = c(
            "South Africa or province" = "province",
            "South Africa or population group" = "population_group"
          ),
          selected = "province"
        )),
        shiny::column(4, shiny::selectizeInput(
          ids[[2L]], "Geography or population group",
          choices = province_choices,
          selected = make_geography_key("national", 10L),
          multiple = FALSE
        )),
        shiny::column(5, shiny::selectInput(
          ids[[3L]], "Cause hierarchy",
          choices = result_hierarchy_choices,
          selected = "all"
        ))
      ),
      shiny::fluidRow(
        shiny::column(6, shiny::selectizeInput(
          ids[[4L]], "Causes or aggregates",
          choices = cause_choices_for_hierarchy("all", grouped = TRUE),
          selected = default_cause_ids("all", multiple = TRUE),
          multiple = TRUE
        )),
        shiny::column(2, shiny::selectizeInput(
          ids[[5L]], "Sex",
          choices = named_choices(sexes, "sex_code", "sex"),
          selected = safe_default(sexes$sex_code, 3L),
          multiple = FALSE
        )),
        shiny::column(2, shiny::selectizeInput(
          ids[[6L]], "Age group",
          choices = named_choices(ages[age_id != "asr_all"], "age_id", "age_label"),
          selected = safe_default(ages$age_id, "age_all"),
          multiple = FALSE
        )),
        shiny::column(2, shiny::selectInput(
          ids[[7L]], "Measure",
          choices = measure_choices,
          selected = "crude_rate"
        ))
      ),
      shiny::sliderInput(
        ids[[8L]], "Years",
        min = as.integer(config$labels$build$start_year),
        max = as.integer(config$labels$build$end_year),
        value = c(
          as.integer(config$labels$build$start_year),
          as.integer(config$labels$build$end_year)
        ),
        sep = ""
      )
    ),
    htmltools::div(
      class = "report-chart wide-report-chart",
      highcharter::highchartOutput(paste0(prefix, "_chart"), height = "540px")
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the selected cause comparison"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download selected cause comparison",
          class = "download-link"
        )
      )
    )
  )
}

register_result_cause_panel <- function(prefix = "result_cause") {
  ids <- list(
    scope = paste0(prefix, "_scope"),
    geography = paste0(prefix, "_geography"),
    hierarchy = paste0(prefix, "_hierarchy"),
    causes = paste0(prefix, "_causes"),
    sex = paste0(prefix, "_sex"),
    age = paste0(prefix, "_age"),
    measure = paste0(prefix, "_measure"),
    years = paste0(prefix, "_years"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  shiny::observeEvent(input[[ids$scope]], {
    scope <- input[[ids$scope]] %||% "province"
    choices <- geography_choices_for_scope(scope)
    selected <- if (identical(scope, "population_group")) {
      unname(choices)[[1L]]
    } else {
      make_geography_key("national", 10L)
    }
    shiny::updateSelectizeInput(
      session, ids$geography,
      choices = choices,
      selected = selected,
      server = TRUE
    )
  }, ignoreInit = TRUE)

  shiny::observeEvent(input[[ids$hierarchy]], {
    hierarchy <- input[[ids$hierarchy]] %||% "all"
    shiny::updateSelectizeInput(
      session, ids$causes,
      choices = cause_choices_for_hierarchy(hierarchy, grouped = identical(hierarchy, "all")),
      selected = default_cause_ids(hierarchy, multiple = TRUE),
      server = TRUE
    )
  }, ignoreInit = TRUE)

  shiny::observeEvent(input[[ids$measure]], {
    update_result_age_choices(ids$age, input[[ids$measure]] %||% "crude_rate")
  }, ignoreInit = FALSE)

  selection_key <- shiny::reactive({
    shiny::req(
      input[[ids$geography]], input[[ids$causes]], input[[ids$sex]],
      input[[ids$age]], input[[ids$measure]], input[[ids$years]]
    )
    list(
      panel = prefix,
      geography = as.character(input[[ids$geography]]),
      causes = as.character(input[[ids$causes]]),
      sex = as.integer(input[[ids$sex]]),
      age = as.character(input[[ids$age]]),
      measure = as.character(input[[ids$measure]]),
      years = as.integer(input[[ids$years]])
    )
  })

  selected_data <- nbd_cached_reactive(
    selection_key,
    function(key) {
      d <- collect_result_slice(
        geography_keys = key$geography,
        sex_code = key$sex,
        series_ids = key$causes,
        age_id = key$age,
        year_range = key$years,
        measure = key$measure
      )
      if (nrow(d)) {
        label_map <- unique(d[, .(series_id, series_label, hierarchy)])
        label_map[, line_name := series_label]
        if (anyDuplicated(label_map$line_name)) {
          label_map[, line_name := paste0(
            series_label, " [", humanise(hierarchy), "]"
          )]
        }
        d <- merge(
          d,
          label_map[, .(series_id, line_name)],
          by = "series_id",
          all.x = TRUE,
          sort = FALSE
        )
        data.table::setorder(d, series_id, year)
      }
      d
    },
    millis = 250L
  )

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 540))
    hc_nbd_lines(
      d,
      y = "estimate",
      series = "line_name",
      lower = "lower",
      upper = "upper",
      title = unique(d$geography)[[1L]],
      subtitle = paste(unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = measure_label(input[[ids$measure]], config),
      config = config,
      decimals = metric_decimals(input[[ids$measure]]),
      height = 540,
      filename = paste0(prefix, "-cause-comparison")
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No NBD3-R result is available for this selection."
      ))
    }
    interval_rows <- d[is.finite(lower) & is.finite(upper), .N]
    htmltools::div(
      class = if (interval_rows) "narrative-box success" else "narrative-box",
      htmltools::strong("Cause comparison. "),
      paste0(
        data.table::uniqueN(d$series_id), " causes or aggregates are displayed; ",
        interval_rows, " of ", nrow(d),
        " plotted cells have exact 95% intervals from ", uncertainty_draw_label, "."
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    measure <- input[[ids$measure]]
    digits <- metric_decimals(measure)
    display <- d[, .(
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Hierarchy = humanise(hierarchy),
      Cause = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      `Draws used` = n_draws
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, measure, digits),
      Lower = format_table_value(Lower, measure, digits),
      Upper = format_table_value(Upper, measure, digits),
      `Draws used` = ifelse(is.na(`Draws used`), "—", format(`Draws used`, big.mark = ","))
    )]
    nbd_kable_html(
      display,
      caption = "Selected NBD3-R causes and aggregates within one population",
      align = c("l", "l", "r", "l", "l", "l", "l", "r", "r", "r", "r")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-cause-comparison.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )
  invisible(TRUE)
}
```

# Study overview and analytical sequence {#overview .unnumbered}

<div class="research-abstract">
<div class="research-abstract-label">Study summary</div>
<p>The Third South African National Burden of Disease Study develops internally coherent estimates of mortality by cause, province, population group, sex, age, and year. It begins with registered cause-of-death microdata and then addresses four central limitations of those data: incomplete death registration, misclassification of HIV/AIDS deaths, incomplete information on the level and composition of injury mortality, and deaths assigned to ill-defined or otherwise unusable underlying causes. The resulting database contains deaths, crude and age-standardised mortality rate. The analytical sequence of the study is shown below:</p>
</div>



<div class="method-flow research-report-flow">
<article><span>1</span><h3>Prepare registered deaths</h3><p>Verify ICD information, map records to NBD causes, aggregate records, and redistribute unknown demographic categories.</p></article>
<article><span>2</span><h3>Complete mortality</h3><p>Place registered deaths within the demographic all-cause envelope and adjust the residual African natural-cause component.</p></article>
<article><span>3</span><h3>Calibrate injuries</h3><p>Use IMS 2009 and FAMHIS 2017 to estimate the injury level and demographic profile, and NIMS/IMS/FAMHIS to estimate cause fractions.</p></article>
<article><span>4</span><h3>Correct HIV/AIDS</h3><p>Estimate the counterfactual non-HIV background and reallocate deaths misclassified to HIV-indicator causes.</p></article>
<article><span>5</span><h3>Redistribute unusable causes</h3><p>Move ill-defined and garbage-coded deaths to the approved targets while preserving mortality envelopes.</p></article>
<article><span>6</span><h3>Construct final outputs</h3><p>Map to the reporting hierarchy, calculate deaths, crude rates and age-standardised rates, and aggregate the complete joint uncertainty draws.</p></article>
</div>


# Background and objectives {#background}

South Africa has long experienced a complex mortality profile in which communicable and poverty-related conditions coexist with non-communicable diseases, injuries, and HIV/AIDS. Registered cause-of-death data provide the central empirical record for analysing this burden, but the raw registrations do not by themselves form a complete or internally coherent account of mortality. Death registration is incomplete in some demographic groups and periods; HIV/AIDS is frequently certified to associated conditions rather than HIV itself; forensic and vital-registration systems provide different information about injury mortality; and a proportion of certificates contain ill-defined, intermediate, or otherwise unusable underlying causes.

The South African NBD programme was established to combine the available demographic and epidemiological evidence into coherent estimates that can support health planning. The initial study produced national estimates for the reference year 2000. The second study extended the analysis to annual national and provincial cause-of-death profiles and formalised the major corrections for under-registration, HIV/AIDS misclassification, injury mortality, and unusable causes. The current analysis translates and extends that analytical system into a reproducible R workflow for 1997–2019.

<div class="research-aims-grid">
<article><span class="overview-eyebrow">Primary objective</span><h3>Estimate coherent mortality levels and causes</h3><p>Produce annual deaths, crude mortality rates, and age-standardised death rates by cause, age, sex, province, and population group.</p></article>
<article><span class="overview-eyebrow">Methodological objective</span><h3>Make every correction explicit and reproducible</h3><p>Retain restartable analytical stages, auditable inputs, documented cause mappings, and transparent conservation checks.</p></article>
<article><span class="overview-eyebrow">Uncertainty objective</span><h3>Propagate linked uncertainty jointly</h3><p>Carry uncertainty through completeness, injury estimation, HIV/AIDS reallocation, and redistribution rather than attaching independent intervals after the analysis.</p></article>
</div>

The specific aims of this report are to document the evidence and statistical operations used in the current analysis; show how the corrections alter the interpretation of registered mortality; compare the NBD3-R estimates with NBD2, the working NBD3-Stata series, THEMBISA, and GBD2023 where suitable comparison data are available; and provide interactive access to the final South African NBD cause hierarchy.

# Data and analytical inputs {#data}

## Core data sources

```{r data-source-table}
data_source_table <- data.table::data.table(
  Source = c(
    "Statistics South Africa vital-registration microdata",
    "Population estimates",
    "All-cause completeness inputs",
    "National Population Register",
    "NIMS 2000",
    "Injury Mortality Survey 2009",
    "FAMHIS 2017",
    "Antenatal HIV prevalence",
    "Cause mappings, YLL schedules, and standard-population weights"
  ),
  `Period or anchor` = c(
    "1997–2019", "1997–2019", "Annual inputs through the NBD2 completeness period",
    "Post-2016 update", "2000", "2009", "2017", "Annual provincial series",
    "Fixed analytical inputs"
  ),
  `Role in the analysis` = c(
    paste(
      "Record-level underlying-cause information; the source for registered",
      "deaths before demographic, completeness, HIV, injury, and garbage-code corrections."
    ),
    paste(
      "Denominators by province, population group, sex, age, and year for",
      "mortality rates, HIV models, and final aggregation."
    ),
    paste(
      "Adjusted all-cause totals and completeness factors developed for the",
      "South African NBD programme by Prof. Rob Dorrington; used to derive",
      "the residual adjustment to African natural deaths."
    ),
    paste(
      "Independent administrative mortality counts used to update the post-NBD2",
      "completeness trajectory after calibration to the overlapping VR period."
    ),
    paste(
      "National injury cause composition by sex and age; geographically expanded",
      "using the IMS distribution."
    ),
    paste(
      "Survey-derived 2009 national and provincial injury level, demographic",
      "profile, well-specified cause fractions, and survey-sampling uncertainty."
    ),
    paste(
      "Later-cleaned 2017 injury level, provincial and demographic profile,",
      "well-specified cause fractions, and survey-sampling uncertainty."
    ),
    paste(
      "Exposure series for the statistical identification of deaths misclassified",
      "from HIV/AIDS to associated causes."
    ),
    paste(
      "Translation from analysis causes to the South African reporting hierarchy,",
      "remaining life expectancy for YLLs, and the standard population for ASRs."
    )
  )
)
nbd_kable_html(
  data_source_table,
  caption = "Principal data sources and their analytical roles",
  align = c("l", "l", "l")
)
```

The vital-registration microdata are the primary cause-of-death source. The complementary inputs are used to correct known limitations of registration rather than to replace the registered series wholesale. The completeness inputs define the demographic mortality boundary; the NPR informs recent changes in registration; injury surveys provide mortality information that is not recoverable from routine coding alone; and antenatal HIV prevalence supports the counterfactual model for HIV/AIDS misclassification.

## Analytical dimensions

```{r analytical-dimensions-table}
analytical_dimensions <- data.table::data.table(
  Dimension = c("Year", "Geography", "Population group", "Sex", "Age", "Cause"),
  Detail = c(
    "Annual estimates, 1997–2019",
    "Nine provinces and South Africa",
    "African, White, Asian/Indian, Coloured, plus combined person totals where defined",
    "Male, female, and person",
    "Detailed analytical age groups and predefined aggregates",
    paste(
      "Analysis causes mapped to all causes, broad groups, disease categories,",
      "detailed causes, and reporting/custom aggregates"
    )
  )
)
nbd_kable_html(
  analytical_dimensions,
  caption = "Dimensions of the analytical database",
  align = c("l", "l")
)
```

# Methods {#methods}

## Preparing the vital-registration data and assigning causes {#vr-preparation}

The pipeline reads the record-level Stats SA mortality data, standardises certificate and demographic fields, and verifies the ICD information before assigning each record to an NBD analysis cause. The cause mapping is supplemented by age, sex, perinatal, cancer, and multiple-cause plausibility rules, together with the documented exclusions and targeted corrections. Records are then aggregated to the most detailed analytical key:

$$
\text{province}\times\text{population group}\times\text{sex}\times
\text{age}\times\text{year}\times\text{cause}.
$$

Three initial demographic redistributions are conducted before the substantive mortality corrections:

1. unknown sex is redistributed within the most detailed observed cause-demographic group;
2. unknown age is redistributed within province, sex, population group, cause, and year; and
3. unknown population group is redistributed using contemporaneous observed shares, with the 1999–2000 reference distribution used for 1997–1998.

Each operation conserves represented deaths. These steps produce a complete demographic-cause table suitable for comparison with the all-cause envelope and the external epidemiological inputs.

## Correcting incomplete death registration {#completeness-methods}

### Residual adjustment to African natural deaths

The completeness inputs specify an adjusted all-cause mortality envelope for each province, sex, age, and year. The registered deaths are divided into four components: African injuries, African natural deaths, injuries in the other population groups, and natural deaths in the other population groups. Let the adjusted all-cause total be $A$ and let the four registered components be $D_{11}$, $D_{12}$, $D_{21}$, and $D_{22}$, where $D_{12}$ is African natural mortality.

When the sum of the components that are not African natural deaths exceeds the adjusted envelope, they are scaled downward using

$$
S_1=\min\left(1,\frac{A}{D_{11}+D_{21}+D_{22}}\right).
$$

The residual scalar for African natural deaths is then

$$
S_2=
\frac{A-S_1(D_{11}+D_{21}+D_{22})}{D_{12}}.
$$

The corrected African natural-cause total is $S_2D_{12}$. This formulation places the residual under-registration adjustment on African natural deaths while preserving the observed injury and other-population components unless the completed envelope requires a common downward adjustment. The neonatal scalar follows the adjacent post-neonatal information where direct estimation is unavailable.

The completeness inputs were developed for the South African NBD programme and are treated as an explicit project input. The broader NBD2 method describes demographic adjustment of under-registration; the residual $S_1/S_2$ implementation is retained in the analytical code and audit tables.

### Updating the recent period with the National Population Register

The original completeness series is held at the configured reference year after 2010. From 2017, the pipeline uses NPR information to update the trajectory. NPR natural and injury counts are aligned to vital registration using their 2015–2016 overlap, after which the calibrated NPR series informs the post-2016 mortality adjustment. The aim is to incorporate recent changes in registration without imposing the older completeness pattern indefinitely.

## Correcting misclassified HIV/AIDS deaths {#hiv-methods}

### Analytical basis

For NBD2, let $D(h,x,t)$ denote completeness-adjusted deaths for cause $h$, age $x$, and time $t$, and let $N(x,t)$ denote population counts. Mortality was related to antenatal HIV prevalence $p(t)$ by age:

$$
\frac{D(x,t)}{N(x,t)} =
\begin{cases}
\alpha(x)+\beta(x)p(t), & x\geq5,\\
\exp\{\alpha(x)+\beta(x)p(t)\}, & x<5.
\end{cases}
$$

A counterfactual background trend was estimated from older-age mortality, and the age-specific baseline was allocated to causes using the 1997 cause shares. Misclassified AIDS deaths were estimated as the excess above the projected non-HIV background.

The current model retains that epidemiological structure while introducing province- and sex-specific background trends, Poisson rather than OLS estimation, triangular age weights, and a combined-sex model below age five. For each HIV-indicator cause,

$$
D(h,x,t)\sim\operatorname{Poisson}\{\lambda(h,x,t)\},\qquad
\log\lambda=\log N+\mu_h+\gamma_h\log(1+t),
$$

with the counterfactual non-HIV rate

$$
R_{\mathrm{non-HIV}}(h,x,t)=
\exp\{\alpha(x,h)+w(x)\gamma_h\log(1+t)\}.
$$

The estimated transfer is

$$
\Omega(h,x,t)=D(h,x,t)-R_{\mathrm{non-HIV}}(h,x,t)N(x,t).
$$

The calculation is run after the completeness and injury-envelope corrections because those operations change the natural-cause mortality available to the HIV model. HIV/AIDS deaths are transferred from the approved indicator causes within the same province × population group × sex × age × year cell, so the cell total is preserved.

### HIV/AIDS estimates across models {#hiv-model-comparison}

The panel below compares the reproducible NBD3-R estimates with the available NBD2, NBD3-Stata, THEMBISA, and GBD2023 series. NBD3-Stata is retained as the internal implementation benchmark, while THEMBISA and GBD provide external comparison trajectories.

```{r register-hiv-comparison, include=FALSE}
register_comparison_panel("hiv_compare", "hiv", "hiv_deaths")
```

```{r hiv-comparison-ui}
comparison_panel_ui("hiv_compare", "hiv", "hiv_deaths")
```

## Correcting the injury envelope and estimating injury causes {#injury-methods}

The injury analysis separates three related quantities:

$$
\text{cause-specific injury deaths}
=
\text{injury level}
\times
\text{demographic profile}
\times
\text{cause fraction}.
$$

All records meeting the published-compatible IMS 2009 and later-cleaned FAMHIS 2017 eligibility rules contribute to the injury level and province–sex–age profile. Cause fractions use only records that map directly to one of the common, well-specified injury causes. Generic and unresolved injury records therefore contribute to the completeness envelope but are not treated as direct evidence for a particular cause.

### Injury evidence sources

<div class="anchor-grid">
<div class="anchor-card"><span class="anchor-year">2000</span><h3>NIMS</h3><p>National cause composition by sex and age. IMS supplies the geographic and population-group distribution used to expand the national NIMS composition to the analytical grid.</p></div>
<div class="anchor-card"><span class="anchor-year">2009</span><h3>IMS</h3><p>Weighted national and provincial injury level, demographic profile, and well-specified cause fractions, with mortuary-level sampling information.</p></div>
<div class="anchor-card"><span class="anchor-year">2017</span><h3>FAMHIS</h3><p>Later-cleaned weighted injury level, forensic-pathology-service profile, and well-specified cause fractions for the second survey anchor.</p></div>
</div>

```{r injury-source-table}
if (nrow(injury_source_table)) {
  nbd_kable_html(
    injury_source_table,
    caption = paste(
      "Observed weighted deaths among records assigned to the common",
      "well-specified injury causes. Each cell gives weighted deaths and",
      "the corresponding share within the displayed cause-fraction evidence."
    ),
    align = c("r", "l", rep("r", ncol(injury_source_table) - 2L)),
    escape = FALSE,
    table_class = "nbd-kable injury-survey-table",
    wrap_class = "nbd-kable-wrap injury-survey-table-wrap"
  )
}
```

### Survey-derived injury level and completeness correction

The empirical weighted survey totals, rather than hard-coded publication values, define the 2009 and 2017 injury anchors. The published estimates remain external validation benchmarks. At survey year $y$, the empirical survey level $S_y$ is compared with the completed routine injury total $V_y$:

$$
K_y=\frac{S_y}{V_y}.
$$

The IMS calibration factor is applied through 2009. For 2010–2016, the log factor is linearly interpolated between IMS and FAMHIS, and the FAMHIS factor is applied from 2017 through 2019. The survey and routine injury totals are also normalised across province × sex × age; their share ratio defines the relative demographic profile. A bounded annual calibration allocates the target injury total across demographic cells while preserving every completed all-cause cell and preventing negative natural mortality.

```{r injury-level-reference-audit}
if (nrow(injury_level_audit)) {
  level_display <- injury_level_audit[, .(
    Survey = survey,
    Year = survey_year,
    `Empirical weighted total` = round(empirical_level_estimate),
    `External reference` = round(reference_estimate),
    `Reference lower` = round(reference_lower),
    `Reference upper` = round(reference_upper),
    `Difference` = round(difference, 1),
    `Difference (%)` = round(100 * relative_difference, 3),
    `Well-specified cause mass (%)` = round(100 * well_specified_share, 2),
    `Reference version` = reference_version
  )]
  nbd_kable_html(
    level_display,
    caption = paste(
      "Survey-derived national injury totals compared with external",
      "publication benchmarks. The empirical totals define the NBD3-R anchors."
    ),
    align = c("l", "r", "r", "r", "r", "r", "r", "r", "r", "l")
  )
}
```

```{r injury-level-anchor-chart}
if (nrow(injury_anchor_data) && all(c(
  "survey_year", "routine_level_total", "survey_level_total"
) %in% names(injury_anchor_data))) {
  anchor_plot <- data.table::rbindlist(list(
    injury_anchor_data[, .(
      year = survey_year,
      series = "Completed VR injury envelope",
      estimate = routine_level_total,
      lower = NA_real_,
      upper = NA_real_
    )],
    injury_anchor_data[, .(
      year = survey_year,
      series = "NBD3-R calibrated survey anchor",
      estimate = survey_level_total,
      lower = NA_real_,
      upper = NA_real_
    )],
    injury_anchor_data[
      is.finite(published_survey_level_total),
      .(
        year = survey_year,
        series = "Published validation benchmark",
        estimate = published_survey_level_total,
        lower = published_survey_level_lower,
        upper = published_survey_level_upper
      )
    ]
  ), use.names = TRUE, fill = TRUE)
  hc_nbd_lines(
    anchor_plot,
    x = "year",
    y = "estimate",
    series = "series",
    lower = "lower",
    upper = "upper",
    title = "Injury mortality at the two survey anchors",
    subtitle = paste(
      "The empirical survey totals define the NBD3-R calibration;",
      "the publication values are external validation benchmarks"
    ),
    x_title = "Survey year",
    y_title = "Injury deaths",
    config = config,
    decimals = 0,
    height = 450,
    filename = "injury-level-survey-anchors",
    markers = TRUE,
    series_order = c(
      "Completed VR injury envelope",
      "NBD3-R calibrated survey anchor",
      "Published validation benchmark"
    )
  )
} else {
  htmltools::div(
    class = "narrative-box warning",
    "The Stage 04 injury-envelope anchor file is not available."
  )
}
```

### Injury cause fractions over time

The 15 cause fractions form a closed composition. Four broad groups are represented first—transport injuries, other unintentional injuries, self-harm, and interpersonal violence—followed by the detailed causes within transport and other unintentional injuries. For non-reference component $h$ and reference component $r$,

$$
z_{hjs}=\log\left(\frac{p_{hjs}}{p_{rjs}}\right).
$$

The log ratios are interpolated between NIMS 2000, IMS 2009, and FAMHIS 2017; the tails are held at the nearest survey anchor; and a five-year triangular moving average with weights $1,2,3,2,1$ reduces survey-year slope changes. Hierarchical inverse-softmax transformations return positive fractions summing to one.

Causes 132, 136, and 138 are harmonised before interpolation because their definitions or coverage are not directly comparable across all three survey instruments. The production anchors use the established cross-survey formulas, after which the complete composition is renormalised. This prevents coding differences from appearing as implausibly steep epidemiological trends.

```{r injury-composition-chart}
output$injury_composition_chart <- highcharter::renderHighchart({
  if (!nrow(injury_broad$model)) {
    return(empty_nbd_chart(
      "The injury model checkpoints are unavailable.",
      height = 500
    ))
  }
  hc_injury_composition(
    injury_broad$source,
    injury_broad$model,
    config,
    title = "National injury composition: survey evidence and annual trajectory",
    subtitle = paste(
      "Cause fractions use well-specified records; total injury completeness",
      "uses all records meeting the published-compatible injury criteria"
    ),
    height = 500
  )
})
highcharter::highchartOutput("injury_composition_chart", height = "500px")
```

```{r injury-broad-survey-model-table}
if (nrow(injury_broad_fit_table)) {
  injury_broad_display <- data.table::copy(injury_broad_fit_table)
  numeric_columns <- names(injury_broad_display)[vapply(
    injury_broad_display, is.numeric, logical(1)
  )]
  injury_broad_display[, (numeric_columns) := lapply(.SD, round, 2),
                       .SDcols = numeric_columns]
  nbd_kable_html(
    injury_broad_display,
    caption = paste(
      "Broad injury composition at the survey anchors compared with the",
      "final annual trajectory on the common weighting basis"
    )
  )
}
```

```{r injury-flagged-survey-model-table}
if (nrow(injury_flagged_fit_table)) {
  injury_flagged_display <- data.table::copy(injury_flagged_fit_table)
  numeric_columns <- names(injury_flagged_display)[vapply(
    injury_flagged_display, is.numeric, logical(1)
  )]
  injury_flagged_display[, (numeric_columns) := lapply(.SD, round, 3),
                         .SDcols = numeric_columns]
  nbd_kable_html(
    injury_flagged_display,
    caption = paste(
      "Survey inputs and final estimates for causes 132, 136, and 138 after",
      "the cross-survey harmonisation rules are incorporated into the trajectory"
    )
  )
}
```

### Injury estimates across models {#injury-model-comparison}

```{r register-injury-comparison, include=FALSE}
register_comparison_panel("injury_compare", "injury", "injuries_deaths")
```

```{r injury-comparison-ui}
comparison_panel_ui("injury_compare", "injury", "injuries_deaths")
```

## Redistributing ill-defined and otherwise unusable causes {#redistribution-methods}

After the completeness, injury, and HIV corrections, the pipeline applies an ordered set of expert rules to deaths that remain assigned to ill-defined symptoms, intermediate causes, mechanisms of death, or insufficiently specified conditions. Each source cause has a defined list of eligible targets. In the deterministic estimate, a source total $S$ is distributed in proportion to the current target counts $D_j$:

$$
A_j=S\frac{D_j}{\sum_kD_k}.
$$

The rules are applied sequentially so that later redistributions use the cause distribution produced by earlier rules. Age, sex, perinatal, and biological restrictions remain in force. When all eligible target counts are zero, the pipeline uses the documented reference distribution rather than dropping the source deaths. The natural, injury, and all-cause envelopes are conserved throughout.

## Quantifying joint uncertainty {#uncertainty-methods}

Uncertainty is propagated by rerunning the linked analytical corrections inside each joint draw. The analysis does not calculate independent standard errors for final causes and then add them after aggregation. Instead, each draw preserves the transfers and covariance generated by the full cause vector.

<div class="method-flow uncertainty-flow">
<article><span>1</span><h3>Completeness</h3><p>Draw one mean-one factor per province, with the log SD estimated from the annual aggregate S2 pattern across time.</p></article>
<article><span>2</span><h3>Injury survey replicate</h3><p>Resample IMS mortuaries and FAMHIS facilities within strata; jointly derive injury level, demographic profile, and well-specified cause counts.</p></article>
<article><span>3</span><h3>HIV/AIDS model</h3><p>Draw fitted coefficient vectors from their variance-covariance matrices and rerun the complete misclassification calculation.</p></article>
<article><span>4</span><h3>Redistribution</h3><p>Draw continuous positive target multipliers for every multi-target rule and rerun the ordered proportional algorithm.</p></article>
<article><span>5</span><h3>Final hierarchy</h3><p>Map each complete draw to the same cause hierarchy, geographies, population groups, ages, and measures used by the point database.</p></article>
</div>

```{r uncertainty-drivers}
uncertainty_drivers <- data.table::data.table(
  Driver = c(
    "African natural-cause completeness",
    "Injury mortality level and demographic profile",
    "Injury cause fractions",
    "HIV/AIDS misclassification",
    "Ill-defined and garbage-code redistribution"
  ),
  `Evidence used` = c(
    paste(
      "Within each province, annual aggregate S2 values after sex and age",
      "are collapsed with implied pre-adjustment African natural deaths as weights."
    ),
    paste(
      "Stratified PSU resampling of IMS mortuaries and FAMHIS forensic",
      "pathology facilities using the supplied weights and strata."
    ),
    paste(
      "The same IMS/FAMHIS survey replicate supplies well-specified cause",
      "counts; NIMS uses count-based uncertainty."
    ),
    paste(
      "The fitted Stage 05 coefficient vectors and their estimated",
      "variance-covariance matrices."
    ),
    paste(
      "The same expert-approved target lists, with continuous positive",
      "multipliers varying the relative target weights in every draw."
    )
  ),
  `Operation in each draw` = c(
    paste(
      "Draw one independent mean-one factor per province and apply it to",
      "African natural deaths before injury and HIV corrections are rerun."
    ),
    paste(
      "Recalculate the 2009 and 2017 survey levels and profiles, interpolate",
      "the survey-to-VR correction, and preserve the completed all-cause cells."
    ),
    paste(
      "Harmonise causes 132, 136, and 138, interpolate the sampled cause",
      "compositions, smooth the path, and apply it to the sampled envelope."
    ),
    paste(
      "Sample coefficients jointly, reconstruct the counterfactual background,",
      "and transfer deaths between HIV/AIDS and its donor causes."
    ),
    paste(
      "Weight the current target counts by the random multiplier vector,",
      "normalise the weights, and redistribute the complete source total."
    )
  )
)
nbd_kable_html(
  uncertainty_drivers,
  caption = "Evidence and stochastic operation for each joint uncertainty driver",
  align = c("l", "l", "l")
)
```

### Completeness uncertainty

For province $p$, the annual aggregate completeness values are formed after age and sex are collapsed. The death-weighted standard deviation of $\log S_{2,pt}$ across time is $\sigma_p$. Each draw uses

$$
F_p^{(b)}=\exp\left\{-\frac{1}{2}\sigma_p^2+\sigma_pZ_p^{(b)}\right\},
\qquad Z_p^{(b)}\sim N(0,1).
$$

The mean-one centring preserves the deterministic point estimate in expectation. The factor is applied to all African natural causes in the province, after which the injury and HIV models are rerun. Cross-province covariance is not imposed because no covariance input is supplied.

### Injury survey uncertainty

Within each survey, one stratified PSU bootstrap replicate jointly determines the national injury level, province–sex–age profile, and well-specified cause counts. This preserves the covariance between the level, geography, age, sex, and cause composition. The same 2009/2017 calibration and interpolation rules used for the point estimate are then applied inside the draw. Total injury intervals therefore retain survey-level uncertainty; specific injury causes additionally retain cause-composition uncertainty.

### HIV/AIDS parameter uncertainty

For each fitted model,

$$
\boldsymbol{\beta}^{(b)}\sim
N\left(\widehat{\boldsymbol{\beta}},
\widehat{\operatorname{Var}}(\widehat{\boldsymbol{\beta}})\right).
$$

The model is rerun after the completeness and injury draws. A draw that reallocates more deaths to HIV/AIDS necessarily retains fewer deaths in the linked donor causes, generating the appropriate negative covariance while preserving the cell total.

### Redistribution uncertainty

For an approved target list $\{1,\ldots,J\}$, the deterministic multiplier vector is $(1,\ldots,1)$. In draw $b$, continuous positive multipliers $W_j^{(b)}$ are sampled from the configured Gamma distribution and normalised. The redistributed amount is

$$
A_j^{(b)}=
S\frac{D_jW_j^{(b)}}{\sum_kD_kW_k^{(b)}}.
$$

Every approved target remains available; no unapproved cause is introduced; and single-target rules remain deterministic. This replaces the small discrete set of target-on/target-off permutations with a continuous distribution over the relative target weights.

# Results {#results}

The results section provides two complementary views of the current NBD cause hierarchy. The first holds the cause fixed and compares the selected aggregation across South Africa, provinces, or population groups. The second holds the population fixed and compares several causes or aggregates on the same plot.

<div class="report-callout callout-info">
<div class="callout-title">Uncertainty coverage in the interactive results</div>
<div class="callout-body">The current joint uncertainty run retains base-age deaths for Male, Female and Person across South Africa, the nine provinces, and the four national population groups. The report derives exact draw-based intervals for every supported cause and aggregate, for deaths and crude rates at the selected age, and for all-age age-standardised rates calculated within each draw. YLLs remain in the analytical database but are not exposed in this visualization tool.</div>
</div>

## Comparing one aggregation across geographies or population groups {#result-geography-comparison}

Use this explorer to select one cause or aggregate, then compare its trend across South Africa and the nine provinces or across the four population groups. The hierarchy control includes all causes, broad groups, disease categories, detailed causes, and reporting/custom aggregates.

```{r register-result-geography, include=FALSE}
register_result_geography_panel("result_geography")
```

```{r result-geography-ui}
result_geography_panel_ui("result_geography")
```

## Comparing causes within one geography or population group {#result-cause-comparison}

Use this explorer to hold the population fixed and display several causes or aggregates together. For example, HIV/AIDS, tuberculosis, and injuries can be compared for South Africa, one province, or one population group under the same sex, age, measure, and year selection.

```{r register-result-cause, include=FALSE}
register_result_cause_panel("result_cause")
```

```{r result-cause-ui}
result_cause_panel_ui("result_cause")
```

## Number of deaths by broad cause group and age group {#result-broad-cause-age}

This view shows how the age distribution of mortality changed across the four broad cause groups. The four panels use a common vertical scale and fixed reference years—1997, 2005, 2012, and 2019—so the changing level and composition of mortality can be compared directly. Select South Africa, one province, or one national population group, together with Male, Female, or Person.

```{r register-result-broad-age, include=FALSE}
register_broad_age_profile_panel("result_broad_age")
```

```{r result-broad-age-ui}
broad_age_profile_panel_ui("result_broad_age")
```

# Discussion and next analytical phases {#discussion}

## Interpretation of the current analysis

The current analysis provides a reproducible 1997–2019 mortality series in which the major data-quality corrections are applied in a fixed sequence and the resulting transfers are preserved through the final cause hierarchy. The all-cause envelope is determined before causes are reallocated. Injury calibration changes the division between injury and natural mortality without adding deaths to the completed total. HIV/AIDS correction transfers deaths from defined donor causes rather than increasing the cell total. Ill-defined redistribution conserves the source total and uses only approved targets. These properties are central to interpreting the results and to understanding why cause-specific uncertainty does not necessarily translate into equally wide all-cause intervals.

The model-comparison panels should be interpreted as diagnostic rather than as a requirement that all series agree. NBD3-Stata provides an internal implementation benchmark; NBD2 shows continuity with the preceding national study; THEMBISA provides a specialised HIV comparison; and GBD2023 provides an external global model. Differences can arise from source data, cause definitions, demographic envelopes, smoothing, and the way uncertainty is represented.

## Extending the mortality series through the pandemic period

The current Version 1 analysis ends in 2019. The proposed Version 2 analysis will add 2020–2022 or 2023, depending on the final data release, and will explicitly quantify COVID-19 mortality. That work requires pandemic-specific treatment of registration delays, excess mortality, cause certification, and redistribution among respiratory, cardiovascular, metabolic, and ill-defined causes. It should be implemented as a new analytical stage rather than by extrapolating the pre-pandemic model unchanged.

## Adding comparative risk assessment

A later phase will link the cause-specific mortality and YLL estimates to comparative risk assessment. This will require harmonised exposure distributions, relative-risk functions, counterfactual exposure distributions, mediation and overlap rules, and uncertainty propagation from both the mortality estimates and the risk model. The mortality database produced here provides the denominator and cause framework on which that work can be built.

## Disaggregating the burden to health districts

A further extension will estimate mortality at health-district level. This will require consistent district coding over time, population denominators, treatment of sparse cause-age-sex cells, and explicit spatial or small-area borrowing. District results should preserve the national and provincial totals while quantifying the additional uncertainty introduced by finer geographic disaggregation.

<div class="future-work-grid">
<article><span>Next</span><h3>2020–2022/23 and COVID-19</h3><p>Extend the cause series through the pandemic and add an explicit COVID-19 correction and uncertainty model.</p></article>
<article><span>Later</span><h3>Comparative risk assessment</h3><p>Estimate the burden attributable to major behavioural, metabolic, environmental, and occupational risks.</p></article>
<article><span>Later</span><h3>Health-district estimates</h3><p>Develop sub-provincial estimates using compatible denominators, geographic harmonisation, and small-area methods.</p></article>
</div>

# Technical appendix {#technical-appendix .unnumbered}

## How uncertainty appears in different results

```{r uncertainty-by-result}
uncertainty_by_result <- data.table::data.table(
  Result = c(
    "All causes", "HIV/AIDS", "Other natural causes",
    "Total injuries", "Specific injury causes", "Population-group causes"
  ),
  `Uncertainty carried into the result` = c(
    paste(
      "General completeness changes the total. Injury calibration, HIV/AIDS",
      "reallocation, cause composition, and redistribution conserve it."
    ),
    paste(
      "Province completeness, fitted HIV/AIDS coefficients, reciprocal donor",
      "changes, and any subsequent eligible redistribution."
    ),
    paste(
      "Completeness, the reciprocal injury-envelope offset in African natural",
      "deaths, HIV/AIDS reallocation where linked, and redistribution."
    ),
    paste(
      "IMS/FAMHIS survey-level and demographic-profile uncertainty; cause",
      "composition cancels when every injury cause is summed."
    ),
    paste(
      "Survey level and profile, NIMS/IMS/FAMHIS cause composition,",
      "harmonisation, interpolation, smoothing, and eligible redistribution."
    ),
    paste(
      "The same complete draw aggregated nationally within each population",
      "group; no proportional allocation of provincial intervals is used."
    )
  ),
  `Changes the all-cause total?` = c(
    "Yes — through completeness",
    "No — reallocation preserves the cell total",
    "Completeness changes the total; other transfers preserve it",
    "No — injury and natural deaths remain complements of the completed total",
    "No — injury and natural deaths remain complements of the completed total",
    paste(
      "The African population group carries the residual completeness and",
      "injury offset; province-level all-cause totals remain conserved"
    )
  )
)
nbd_kable_html(
  uncertainty_by_result,
  caption = "How the joint uncertainty components appear in reported outcomes",
  align = c("l", "l", "l")
)
```

## Aggregation and covariance

The final hierarchy is constructed separately within every draw and only then summarised. For all-cause mortality,

$$
T^{(b)}=\sum_cD_c^{(b)},
$$

so

$$
\operatorname{Var}(T)=
\sum_c\operatorname{Var}(D_c)+
2\sum_{c<k}\operatorname{Cov}(D_c,D_k).
$$

The covariance term is essential. HIV/AIDS and its donor causes can move in opposite directions; injury causes compete within a closed composition; and redistribution transfers a fixed source total among eligible recipients. The report therefore does not construct an aggregate interval by adding independent cause variances.

## Current uncertainty boundaries

```{r uncertainty-limitations}
uncertainty_limitations <- data.table::data.table(
  Quantity = c(
    "Monte Carlo draw count", "S1 completeness", "Cross-province completeness covariance",
    "Independent NPR variance", "Antenatal HIV prevalence",
    "Injury survey design", "Reported measures"
  ),
  `Current treatment` = c(
    paste(
      uncertainty_draw_label,
      "in the configured run; a larger final draw count will stabilise interval tails"
    ),
    "Fixed because no separate sampling-variance input is supplied",
    "Not supplied; the province-specific completeness factors are drawn independently",
    "Fixed because no independent NPR variance source is supplied",
    "Fixed because no approved prevalence variance-covariance input is supplied",
    paste(
      "IMS and FAMHIS use a stratified PSU bootstrap from the supplied facility,",
      "stratum, and weight fields; exact replication of another software's",
      "finite-population and variance settings is not imposed"
    ),
    paste(
      "Current intervals are generated for supported deaths and crude rates;",
      "YLLs remain in the analytical database but are not exposed in this visualization tool"
    )
  )
)
nbd_kable_html(
  uncertainty_limitations,
  caption = "Current boundaries of the uncertainty analysis",
  align = c("l", "l")
)
```

## Study references

Bradshaw D, Groenewald P, Laubscher R, et al. *Initial burden of disease estimates for South Africa, 2000.* South African Medical Journal. 2003;93:682–688.

Matzopoulos R, Prinsloo M, Bradshaw D, et al. *The Injury Mortality Survey: A national study of injury mortality levels and causes in South Africa in 2009.* Cape Town: South African Medical Research Council; 2013.

Msemburi W, Pillay-van Wyk V, Dorrington RE, et al. *Second national burden of disease study for South Africa: Cause-of-death profile for South Africa, 1997–2012.* Cape Town: South African Medical Research Council; 2016.

Awotiwon OF, Pillay-van Wyk V, Groenewald P, et al. *SA NBD–GBD and SA NBD–WHO cause-list mappings for the second South African National Burden of Disease.* Cape Town: South African Medical Research Council; 2017.

Prinsloo M, Mhlongo S, Dekel B, et al. *The 2nd Injury Mortality Survey: A national study of injury mortality levels and causes in South Africa in 2017.* Cape Town: South African Medical Research Council; 2021.

Prinsloo M, Mhlongo S, Roomaney RA, et al. *Injury mortality in South Africa: a 2009 and 2017 comparison to track progress to meeting sustainable development goal targets.* Global Health Action. 2024;17(1):2377828.

<div class="report-footer"><div><strong>Third South African National Burden of Disease Study</strong><br/>Comprehensive analytical report, 1997–2019</div><div>`r config$labels$project$analysis_period`</div></div>
