---
title: "Third South African National Burden of Disease Study"
subtitle: "Version 1: Reproducible mortality estimates for South Africa, 1997–2019"
author: "NBD3 analytical team"
date: "`r format(Sys.Date(), '%d %B %Y')`"
runtime: shiny
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    number_sections: false
    theme: flatly
    highlight: textmate
    fig_caption: true
    css: www/report.css
    df_print: default
    self_contained: true
---

```{r setup, include=FALSE}
options(
  stringsAsFactors = FALSE,
  scipen = 999,
  shiny.sanitize.errors = TRUE
)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

required_packages <- c(
  "arrow", "data.table", "dplyr", "highcharter", "htmltools",
  "knitr", "rmarkdown", "scales", "shiny", "yaml"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop(
    "Missing report package(s): ", paste(missing_packages, collapse = ", "),
    ". Run source('install_packages.R') from the repository root."
  )
}

find_report_root <- function(
    start = dirname(normalizePath(knitr::current_input(), mustWork = TRUE))) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(current, "_targets.R")) &&
        file.exists(file.path(current, "report", "R", "report_data.R"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) stop("Could not locate the NBD repository root.")
    current <- parent
  }
}

root <- find_report_root()
knitr::opts_knit$set(root.dir = root)
source(file.path(root, "report", "R", "report_data.R"))
source(file.path(root, "report", "R", "report_charts.R"))

config <- read_viz_config(root)
runtime <- load_viz_runtime_data(config, open_rates = TRUE)
comparison_data <- data.table::as.data.table(data.table::copy(runtime$comparisons))
cause_uncertainty <- data.table::as.data.table(data.table::copy(
  runtime$cause_uncertainty %||% data.table::data.table()
))
catalog <- runtime$catalogs
injury_inputs <- load_injury_report_inputs(config)
injury_broad <- summarise_injury_broad_fractions(injury_inputs)
injury_broad_comparison <- data.table::as.data.table(data.table::copy(
  injury_broad$comparison
))
injury_detailed_comparison <- data.table::as.data.table(data.table::copy(
  injury_broad$detailed_comparison
))
injury_level_audit_path <- file.path(
  root, "output", "tables", "03_injury_survey_reference_audit.csv"
)
injury_level_audit <- if (file.exists(injury_level_audit_path)) {
  data.table::fread(injury_level_audit_path)
} else {
  data.table::data.table()
}

uncertainty_draw_n <- if (nrow(cause_uncertainty) && "n_draws" %in% names(cause_uncertainty)) {
  values <- cause_uncertainty$n_draws[is.finite(cause_uncertainty$n_draws)]
  if (length(values)) max(values) else NA_integer_
} else {
  NA_integer_
}
uncertainty_draw_label <- if (length(uncertainty_draw_n) == 1L && !is.na(uncertainty_draw_n) && is.finite(uncertainty_draw_n)) {
  paste0(format(uncertainty_draw_n, big.mark = ",", scientific = FALSE), " joint draws")
} else {
  "joint draws"
}
model_names_available <- ordered_model_names(unique(comparison_data$model), config)

measure_choices <- c(
  "Deaths" = "deaths",
  "Crude mortality rate per 100,000" = "crude_rate",
  "Age-standardised rate per 100,000" = "asr",
  "YLLs (undiscounted)" = "yll00",
  "YLLs (3% discount)" = "yll03",
  "YLLs (1.5% discount)" = "yll015"
)

format_value <- function(x, measure, digits = 1L) {
  if (!length(x) || is.na(x) || !is.finite(x)) return("—")
  if (identical(measure, "fraction")) {
    return(scales::percent(x, accuracy = 0.1))
  }
  format(round(x, digits), big.mark = ",", scientific = FALSE, trim = TRUE)
}

metric_decimals <- function(measure) {
  if (measure %in% c("deaths", "yll00", "yll03", "yll015")) 0 else 2
}

safe_default <- function(values, preferred = NULL) {
  values <- as.character(values)
  values <- values[!is.na(values) & nzchar(values)]
  if (!length(values)) return(character())
  if (!is.null(preferred) && as.character(preferred) %in% values) {
    return(as.character(preferred))
  }
  values[[1L]]
}

named_choices <- function(data, value, label) {
  data <- data.table::as.data.table(data.table::copy(data))
  if (!all(c(value, label) %in% names(data))) return(stats::setNames(character(), character()))
  data <- unique(data[, c(value, label), with = FALSE])
  keep <- !is.na(data[[value]]) & !is.na(data[[label]]) &
    nzchar(as.character(data[[value]])) & nzchar(as.character(data[[label]]))
  data <- data[keep]
  stats::setNames(as.character(data[[value]]), as.character(data[[label]]))
}

humanise <- function(x) tools::toTitleCase(gsub("_", " ", as.character(x)))
format_count <- function(x) {
  if (!length(x) || is.na(x) || !is.finite(x)) return("—")
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_table_value <- function(x, measure = NULL, digits = 1L) {
  vapply(
    x,
    function(value) {
      if (is.na(value) || !is.finite(value)) return("—")
      if (!is.null(measure) && identical(measure, "fraction")) {
        return(scales::percent(value, accuracy = 0.1))
      }
      format(
        round(value, digits),
        big.mark = ",",
        scientific = FALSE,
        trim = TRUE,
        nsmall = if (digits > 0L) digits else 0L
      )
    },
    character(1)
  )
}

nbd_kable_html <- function(
    data,
    caption = NULL,
    align = NULL,
    escape = TRUE,
    table_class = "nbd-kable",
    wrap_class = "nbd-kable-wrap",
    empty_message = "No data are available for this selection.") {
  x <- as.data.frame(data, stringsAsFactors = FALSE, check.names = FALSE)
  if (!nrow(x)) {
    x <- data.frame(Note = empty_message, check.names = FALSE)
  }
  if (is.null(align)) {
    align <- rep("l", ncol(x))
  }
  table_html <- knitr::kable(
    x,
    format = "html",
    escape = escape,
    align = align,
    caption = caption,
    table.attr = paste0('class="', table_class, '"')
  )
  htmltools::HTML(paste0(
    '<div class="', wrap_class, '">',
    as.character(table_html),
    '</div>'
  ))
}

comparison_panel_ui <- function(prefix, domain_name, default_series) {
  available <- comparison_data[get("domain") == domain_name]
  geographies <- unique(available[, .(
    geography_type, geography_code, geography
  )])[order(geography_type, geography_code)]
  sexes <- unique(available[, .(sex_code, sex)])[order(sex_code)]
  ages <- unique(available[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  series <- unique(available[, .(
    series_id, series_label, series_sort_order
  )])[order(series_sort_order)]

  ids <- paste0(prefix, c("_geography", "_sex", "_age", "_series"))
  shiny::tagList(
    shiny::fluidRow(
      shiny::column(
        3,
        htmltools::div(
          class = "control-panel",
          shiny::selectizeInput(
            ids[[1L]], "Geography",
            choices = named_choices(geographies, "geography_code", "geography"),
            selected = safe_default(geographies$geography_code, 10L),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[2L]], "Sex",
            choices = named_choices(sexes, "sex_code", "sex"),
            selected = safe_default(sexes$sex_code, 3L),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[3L]], "Age group",
            choices = named_choices(ages, "age_id", "age_label"),
            selected = safe_default(ages$age_id, "age_all"),
            multiple = FALSE
          ),
          shiny::selectizeInput(
            ids[[4L]], "Series",
            choices = named_choices(series, "series_id", "series_label"),
            selected = safe_default(series$series_id, default_series),
            multiple = FALSE
          ),
          htmltools::p(
            class = "control-hint",
            "All available models are shown together. Click a legend item to hide or restore a series."
          )
        )
      ),
      shiny::column(
        9,
        htmltools::div(
          class = "report-chart",
          highcharter::highchartOutput(paste0(prefix, "_chart"), height = "500px")
        )
      )
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the annual model values"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download comparison data",
          class = "download-link"
        )
      )
    )
  )
}

register_comparison_panel <- function(prefix, domain_name, default_series) {
  ids <- list(
    geography = paste0(prefix, "_geography"),
    sex = paste0(prefix, "_sex"),
    age = paste0(prefix, "_age"),
    series = paste0(prefix, "_series"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  selected_data <- shiny::reactive({
    shiny::req(
      input[[ids$geography]], input[[ids$sex]],
      input[[ids$age]], input[[ids$series]]
    )
    comparison_data[
      get("domain") == domain_name &
        geography_code == as.integer(input[[ids$geography]]) &
        sex_code == as.integer(input[[ids$sex]]) &
        age_id == input[[ids$age]] &
        series_id == input[[ids$series]]
    ][order(
      year,
      factor(
        model,
        levels = as.character(unlist(config$labels$models$order, use.names = FALSE))
      )
    )]
  })

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 500))
    selected_measure <- unique(d$measure)[[1L]]
    d[, `:=`(
      plot_estimate = estimate,
      plot_lower = lower,
      plot_upper = upper
    )]
    suffix <- ""
    decimals <- metric_decimals(selected_measure)
    y_title <- measure_label(selected_measure, config)
    if (identical(selected_measure, "fraction")) {
      d[, `:=`(
        plot_estimate = 100 * plot_estimate,
        plot_lower = 100 * plot_lower,
        plot_upper = 100 * plot_upper
      )]
      suffix <- "%"
      decimals <- 1
      y_title <- "Percent of all-cause deaths"
    }
    hc_nbd_lines(
      d,
      y = "plot_estimate",
      series = "model",
      lower = "plot_lower",
      upper = "plot_upper",
      title = unique(d$series_label)[[1L]],
      subtitle = paste(unique(d$geography), unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = y_title,
      config = config,
      decimals = decimals,
      suffix = suffix,
      height = 500,
      filename = paste0(prefix, "-model-comparison"),
      series_order = as.character(unlist(config$labels$models$order, use.names = FALSE))
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No model series are available for the selected cell."
      ))
    }
    model_names <- ordered_model_names(unique(d$model), config)
    interval_models <- ordered_model_names(
      unique(d[is.finite(lower) & is.finite(upper), model]),
      config
    )
    interval_text <- if (length(interval_models)) {
      paste0(
        "Uncertainty bands are available for: ",
        paste(interval_models, collapse = ", "), "."
      )
    } else {
      "No uncertainty limits are supplied for this exact selection."
    }
    htmltools::div(
      class = "narrative-box",
      htmltools::strong("Models shown: "),
      paste(model_names, collapse = ", "),
      htmltools::br(),
      interval_text,
      htmltools::br(),
      htmltools::span(
        class = "provisional-note",
        paste0(
          "NBD3-R intervals come from ", uncertainty_draw_label,
          "; external intervals are retained as supplied."
        )
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    selected_measure <- unique(d$measure)[[1L]]
    digits <- if (identical(selected_measure, "fraction")) {
      1L
    } else {
      metric_decimals(selected_measure)
    }
    display <- d[, .(
      Model = model,
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Series = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      Source = source
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, selected_measure, digits),
      Lower = format_table_value(Lower, selected_measure, digits),
      Upper = format_table_value(Upper, selected_measure, digits)
    )]
    nbd_kable_html(
      display,
      caption = "Annual estimates and uncertainty limits for the selected model comparison",
      align = c("l", "l", "l", "r", "l", "l", "l", "r", "r", "r", "l")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-model-comparison.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )

  invisible(TRUE)
}

final_panel_ui <- function(prefix, geography_type_name, domain_name, default_series) {
  availability <- unique(catalog$availability[
    model == "NBD3-R" &
      get("geography_type") %in% geography_type_name &
      get("domain") == domain_name
  ])
  geographies <- unique(availability[, .(
    geography_code, geography
  )])[order(geography_code)]
  sexes <- unique(availability[, .(sex_code, sex)])[order(sex_code)]
  ages <- unique(availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]
  causes <- unique(availability[, .(
    series_id, series_label, hierarchy, series_sort_order
  )])[order(series_sort_order)]
  grouped_causes <- lapply(
    split(causes, causes$hierarchy),
    function(x) named_choices(x, "series_id", "series_label")
  )
  names(grouped_causes) <- humanise(names(grouped_causes))

  ids <- paste0(prefix, c(
    "_geographies", "_sex", "_cause", "_age", "_measure", "_years"
  ))
  shiny::tagList(
    htmltools::div(
      class = "control-panel final-control-panel",
      shiny::fluidRow(
        shiny::column(4, shiny::selectizeInput(
          ids[[1L]], "Geographies",
          choices = named_choices(geographies, "geography_code", "geography"),
          selected = as.character(geographies$geography_code),
          multiple = TRUE
        )),
        shiny::column(2, shiny::selectizeInput(
          ids[[2L]], "Sex",
          choices = named_choices(sexes, "sex_code", "sex"),
          selected = safe_default(sexes$sex_code, 3L),
          multiple = FALSE
        )),
        shiny::column(6, shiny::selectizeInput(
          ids[[3L]], "Cause",
          choices = grouped_causes,
          selected = safe_default(causes$series_id, default_series),
          multiple = FALSE
        ))
      ),
      shiny::fluidRow(
        shiny::column(4, shiny::selectizeInput(
          ids[[4L]], "Age group",
          choices = named_choices(ages[age_id != "asr_all"], "age_id", "age_label"),
          selected = safe_default(ages$age_id, "age_all"),
          multiple = FALSE
        )),
        shiny::column(4, shiny::selectInput(
          ids[[5L]], "Measure",
          choices = measure_choices,
          selected = "crude_rate"
        )),
        shiny::column(4, shiny::sliderInput(
          ids[[6L]], "Years",
          min = as.integer(config$labels$build$start_year),
          max = as.integer(config$labels$build$end_year),
          value = c(
            as.integer(config$labels$build$start_year),
            as.integer(config$labels$build$end_year)
          ),
          sep = ""
        ))
      )
    ),
    htmltools::div(
      class = "report-chart wide-report-chart",
      highcharter::highchartOutput(paste0(prefix, "_chart"), height = "520px")
    ),
    shiny::uiOutput(paste0(prefix, "_note")),
    htmltools::tags$details(
      class = "appendix-detail",
      htmltools::tags$summary("View and download the selected NBD3 Version 1 values"),
      htmltools::div(
        class = "report-table",
        shiny::uiOutput(paste0(prefix, "_table")),
        shiny::downloadLink(
          paste0(prefix, "_download"),
          "Download selected NBD3 Version 1 data",
          class = "download-link"
        )
      )
    )
  )
}

register_final_panel <- function(prefix, geography_type_name, domain_name, default_series) {
  ids <- list(
    geographies = paste0(prefix, "_geographies"),
    sex = paste0(prefix, "_sex"),
    cause = paste0(prefix, "_cause"),
    age = paste0(prefix, "_age"),
    measure = paste0(prefix, "_measure"),
    years = paste0(prefix, "_years"),
    chart = paste0(prefix, "_chart"),
    note = paste0(prefix, "_note"),
    table = paste0(prefix, "_table"),
    download = paste0(prefix, "_download")
  )

  availability <- unique(catalog$availability[
    model == "NBD3-R" &
      get("geography_type") %in% geography_type_name &
      get("domain") == domain_name
  ])
  age_catalog <- unique(availability[, .(
    age_id, age_label, age_sort_order
  )])[order(age_sort_order)]

  shiny::observeEvent(input[[ids$measure]], {
    selected_measure <- input[[ids$measure]] %||% "crude_rate"
    choices <- if (identical(selected_measure, "asr")) {
      age_catalog[age_id == "asr_all"]
    } else {
      age_catalog[age_id != "asr_all"]
    }
    preferred <- if (identical(selected_measure, "asr")) "asr_all" else "age_all"
    shiny::updateSelectizeInput(
      session,
      ids$age,
      choices = named_choices(choices, "age_id", "age_label"),
      selected = safe_default(choices$age_id, preferred),
      server = TRUE
    )
  }, ignoreInit = FALSE)

  selected_data <- shiny::reactive({
    shiny::req(
      input[[ids$geographies]], input[[ids$sex]], input[[ids$cause]],
      input[[ids$age]], input[[ids$measure]], input[[ids$years]]
    )
    selected_geography_codes <- suppressWarnings(as.integer(unlist(
      input[[ids$geographies]], recursive = TRUE, use.names = FALSE
    )))
    selected_geography_codes <- unique(selected_geography_codes[!is.na(selected_geography_codes)])
    selected_sex_code <- suppressWarnings(as.integer(unlist(
      input[[ids$sex]], recursive = TRUE, use.names = FALSE
    )))[[1L]]
    selected_years <- suppressWarnings(as.integer(unlist(
      input[[ids$years]], recursive = TRUE, use.names = FALSE
    )))
    selected_years <- selected_years[!is.na(selected_years)]

    d <- collect_cause_rates(
      runtime$cause_rates_dataset,
      models = "NBD3-R",
      geography_type = geography_type_name,
      geography_codes = selected_geography_codes,
      sex_code = selected_sex_code,
      series_id = input[[ids$cause]],
      age_id = input[[ids$age]],
      year_range = selected_years,
      measure = input[[ids$measure]]
    )
    if (!nrow(d)) return(d)
    d[, `:=`(
      lower = NA_real_,
      upper = NA_real_,
      n_draws = NA_integer_,
      uncertainty_source = NA_character_
    )]
    if (nrow(cause_uncertainty)) {
      key <- c(
        "geography_type", "geography_code", "sex_code", "year",
        "age_id", "series_id", "measure"
      )
      uncertainty_slice <- cause_uncertainty[
        get("geography_type") %in% geography_type_name &
          geography_code %in% selected_geography_codes &
          sex_code == selected_sex_code &
          series_id == input[[ids$cause]] &
          age_id == input[[ids$age]] &
          measure == input[[ids$measure]] &
          year >= min(selected_years) &
          year <= max(selected_years),
        c(key, "lower", "upper", "n_draws", "source"),
        with = FALSE
      ]
      if (nrow(uncertainty_slice)) {
        data.table::setnames(uncertainty_slice, "source", "uncertainty_source")
        d[, c("lower", "upper", "n_draws", "uncertainty_source") := NULL]
        d <- merge(d, uncertainty_slice, by = key, all.x = TRUE, sort = FALSE)
      }
    }
    d[, line_name := geography]
    d[order(geography_code, year)]
  })

  output[[ids$chart]] <- highcharter::renderHighchart({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(empty_nbd_chart(height = 520))
    hc_nbd_lines(
      d,
      y = "estimate",
      series = "line_name",
      lower = "lower",
      upper = "upper",
      title = unique(d$series_label)[[1L]],
      subtitle = paste(unique(d$sex), unique(d$age_label), sep = " — "),
      x_title = "Year",
      y_title = measure_label(input[[ids$measure]], config),
      config = config,
      decimals = metric_decimals(input[[ids$measure]]),
      height = 520,
      filename = paste0(prefix, "-nbd3-r-trend")
    )
  })

  output[[ids$note]] <- shiny::renderUI({
    d <- selected_data()
    if (!nrow(d)) {
      return(htmltools::div(
        class = "narrative-box warning",
        "No NBD3 Version 1 R results are available for this selection."
      ))
    }
    interval_rows <- d[is.finite(lower) & is.finite(upper), .N]
    total_rows <- nrow(d)
    interval_message <- if (interval_rows > 0L) {
      paste0(
        format(interval_rows, big.mark = ","), " of ",
        format(total_rows, big.mark = ","),
        " plotted cells have 95% uncertainty intervals from ",
        uncertainty_draw_label, "."
      )
    } else if (identical(geography_type_name, "population_group")) {
      paste(
        "The loaded uncertainty profile does not contain an exact interval",
        "for this population-group selection. The Version 1 joint uncertainty output",
        "retains national population-group draws for Person estimates."
      )
    } else if (as.integer(input[[ids$sex]]) != 3L) {
      "The completed uncertainty reporting profile stores Person estimates only; sex-specific results are point estimates."
    } else if (!input[[ids$measure]] %in% c("deaths", "crude_rate")) {
      "The completed draw outputs support deaths and crude rates; ASRs and YLLs are shown as point estimates."
    } else {
      "This cause or age group is outside the exact cause catalogue retained in the completed compact draw outputs."
    }
    htmltools::div(
      class = if (interval_rows > 0L) "narrative-box success" else "narrative-box",
      htmltools::strong("NBD3 Version 1 — R implementation. "),
      interval_message,
      htmltools::br(),
      htmltools::span(
        class = "provisional-note",
        paste(
          "Intervals combine under-reporting, injury-envelope, injury",
          "survey-composition, HIV fitted-parameter, and redistribution uncertainty in one joint run."
        )
      )
    )
  })

  output[[ids$table]] <- shiny::renderUI({
    d <- data.table::copy(selected_data())
    if (!nrow(d)) return(nbd_kable_html(data.frame()))
    selected_measure <- input[[ids$measure]]
    digits <- metric_decimals(selected_measure)
    display <- d[, .(
      Geography = geography,
      Sex = sex,
      Year = year,
      Age = age_label,
      Cause = series_label,
      Measure = measure,
      Estimate = estimate,
      Lower = lower,
      Upper = upper,
      `Draws used` = n_draws
    )]
    display[, `:=`(
      Estimate = format_table_value(Estimate, selected_measure, digits),
      Lower = format_table_value(Lower, selected_measure, digits),
      Upper = format_table_value(Upper, selected_measure, digits),
      `Draws used` = vapply(
        `Draws used`,
        function(value) if (is.na(value)) "—" else format(value, big.mark = ","),
        character(1)
      )
    )]
    nbd_kable_html(
      display,
      caption = "Selected NBD3 Version 1 estimates and 95% uncertainty intervals",
      align = c("l", "l", "r", "l", "l", "l", "r", "r", "r", "r")
    )
  })

  output[[ids$download]] <- shiny::downloadHandler(
    filename = function() paste0(prefix, "-nbd3-r.csv"),
    content = function(file) data.table::fwrite(selected_data(), file)
  )

  invisible(TRUE)
}

# Injury source table used in the methods section.
injury_source_table <- data.table::data.table()
source_fractions <- data.table::as.data.table(data.table::copy(
  injury_inputs$source_fractions
))
if (nrow(source_fractions) && all(c(
  "survey", "year", "nbdcode", "observed_count"
) %in% names(source_fractions))) {
  labels <- data.table::fread(file.path(root, "data", "lookups", "analysis_codes.csv"))
  injury_source_table <- source_fractions[, .(
    observed_deaths = sum(observed_count, na.rm = TRUE)
  ), by = .(survey, year, nbdcode)]
  injury_source_table[, percent := safe_ratio(
    observed_deaths, sum(observed_deaths), 100
  ), by = .(survey, year)]
  injury_source_table <- merge(
    injury_source_table,
    labels[, .(nbdcode = as.integer(analysis_code), Cause = analysis_label)],
    by = "nbdcode",
    all.x = TRUE,
    sort = FALSE
  )
  injury_source_table[is.na(Cause), Cause := paste("Cause", nbdcode)]
  injury_source_table[, display := paste0(
    "<span class=\"injury-table-count\">",
    format(round(observed_deaths), big.mark = ",", scientific = FALSE),
    "</span><span class=\"injury-table-share\">",
    sprintf("%.2f%%", percent),
    "</span>"
  )]
  injury_source_table <- data.table::dcast(
    injury_source_table,
    nbdcode + Cause ~ survey,
    value.var = "display",
    fill = "—"
  )
  data.table::setorder(injury_source_table, nbdcode)
  data.table::setnames(
    injury_source_table,
    c("nbdcode", "Cause"),
    c("Code", "Cause of injury")
  )
  survey_columns <- intersect(
    c("NIMS 2000", "IMS 2009", "FAMHIS 2017"),
    names(injury_source_table)
  )
  data.table::setcolorder(
    injury_source_table,
    c("Code", "Cause of injury", survey_columns)
  )
}

injury_cause_labels <- data.table::fread(
  file.path(root, "data", "lookups", "analysis_codes.csv")
)[, .(
  nbdcode = as.integer(analysis_code),
  Cause = as.character(analysis_label)
)]

injury_broad_fit_table <- data.table::data.table()
if (nrow(injury_broad_comparison)) {
  injury_broad_fit_table <- injury_broad_comparison[, .(
    Survey = survey,
    Year = year,
    `Broad injury group` = broad_group,
    `Survey input (%)` = 100 * survey_fraction,
    `Final smoothed estimate (%)` = 100 * model_fraction,
    `Difference (percentage points)` = difference_percentage_points
  )]
  data.table::setorderv(
    injury_broad_fit_table,
    c("Year", "Broad injury group")
  )
}

injury_flagged_fit_table <- data.table::data.table()
if (nrow(injury_detailed_comparison)) {
  injury_flagged_fit_table <- merge(
    injury_detailed_comparison[nbdcode %in% c(132L, 136L, 138L)],
    injury_cause_labels,
    by = "nbdcode",
    all.x = TRUE,
    sort = FALSE
  )[, .(
    Survey = survey,
    Year = year,
    Code = nbdcode,
    Cause,
    `Survey input (%)` = 100 * survey_fraction,
    `Final smoothed estimate (%)` = 100 * model_fraction,
    `Difference (percentage points)` = difference_percentage_points
  )]
  data.table::setorder(injury_flagged_fit_table, Year, Code)
}
```

<div class="nbd-report-header collaborator-header">
<span class="nbd-report-kicker">Version 1 analytical report</span>
<h1>Third South African National Burden of Disease Study</h1>
<p>This release provides the reproducible R implementation of NBD3 for 1997–2019. It uses the unpublished NBD3 Stata implementation as the internal replication benchmark, carries forward the analytical foundations of NBD2, incorporates the expanded injury evidence base, and introduces formal propagation of analytical uncertainty.</p>
</div>

<div class="utility-note"><strong>Interactivity.</strong> Use the controls embedded in each results section. Charts support zooming, hover values, image export, and data download; the floating navigation panel follows the report as you scroll.</div>

# Overview {#overview}

<div class="section-lead">A national burden of disease study develops consistent and coherent estimates of the levels and underlying causes of death using the best available epidemiological and demographic information. For South Africa this requires addressing deficiencies in the registered cause-of-death data and in particular, accounting for under-registration, correcting for HIV/AIDS misclassification, reallocating undetermined injury causes, as well as reclassifying ill-defined or otherwise unusable underlying causes.</div>

## Development of the South African NBD studies {#study-lineage}

<div class="study-lineage-grid">
<article class="study-generation-card generation-one"><span class="study-generation-label">Initial South African NBD</span><div class="study-generation-period">Reference year 2000</div><h3>National modelling foundation</h3><p>The first study was implemented through spreadsheet-based workflows. Total mortality and the HIV/AIDS epidemic were modelled using ASSA2000; the non-AIDS cause profile drew on registered mortality, Home Affairs data, and national injury surveillance. Its principal focus was premature mortality measured through YLLs, with initial YLD and DALY estimates.</p></article>
<article class="study-generation-card generation-two"><span class="study-generation-label">Second South African NBD</span><div class="study-generation-period">1997–2012</div><h3>National and provincial time series</h3><p>NBD2 moved the analysis into Stata and produced annual national and provincial mortality profiles. It adjusted for incomplete registration, reallocated misclassified HIV/AIDS deaths, estimated specific injury causes from survey evidence, redistributed garbage and ill-defined causes, and reported coherent cause-specific deaths and YLLs.</p></article>
<article class="study-generation-card generation-three"><span class="study-generation-label">Third South African NBD</span><div class="study-generation-period">Target: 1997–2022</div><h3>Reproducible, uncertainty-aware R system</h3><p>NBD3 is being built as a fully scripted R workflow. Version 1 covers 1997–2019 and reproduces some of the working unpublished NBD3 Stata analysis while incorporating NIMS 2000, IMS 2009, and FAMHIS 2017 in the injury model and quantifying some of the uncertainty. The planned extension will add 2020–2022 and explicitly estimate COVID-19 mortality as well.</p></article>
</div>

<div class="report-callout callout-info study-scope-callout">
<div class="callout-title">How to interpret the study and model labels</div>
<div class="callout-body"><strong>NBD3-Stata</strong> and <strong>NBD3-R</strong> are two implementations of the third National Burden of Disease Study. NBD3-Stata is the working model to date; NBD3-R is the updated reproducible implementation used for Version 1 cause estimates. <strong>NBD2</strong> is the preceding national study, while <strong>THEMBISA</strong> and <strong>GBD2023</strong> are external comparison series.</div>
</div>

<div class="nbd-overview-grid">
<article class="overview-panel"><span class="overview-eyebrow">Version 1 scope</span><h3>Annual mortality estimates, 1997–2019</h3><p>Estimate deaths and premature mortality by year, age, sex, province or population group, and cause. Years 2020–2022 and COVID-19 mortality are outside the present release and form the next extension of NBD3.</p></article>
<article class="overview-panel"><span class="overview-eyebrow">Core evidence</span><h3>Combine complementary data sources</h3><p>Use registered cause-of-death records and population denominators together with NIMS 2000, IMS 2009, FAMHIS 2017, completeness estimates, National Population Register information, antenatal HIV prevalence, YLL schedules, and standard-population weights.</p></article>
<article class="overview-panel"><span class="overview-eyebrow">Six-step analytical sequence</span><h3>Correct data problems while preserving the mortality envelope</h3><p>Clean and aggregate registered deaths; complete African natural mortality; calibrate the national injury level with weighted IMS 2009 and FAMHIS 2017 totals and distribute it with their relative province–sex–age profiles; estimate misclassified HIV/AIDS deaths; divide the calibrated injury envelope using the NIMS/IMS/FAMHIS cause profile; and redistribute ill-defined and garbage causes using the approved target rules.</p></article>
<article class="overview-panel"><span class="overview-eyebrow">What this report provides</span><h3>Methods, comparisons, final estimates, and uncertainty</h3><p>Follow the established visualization order, compare the two NBD3 implementations with NBD2 and external benchmarks, and present the final Version 1 cause estimates from the R implementation with uncertainty wherever the completed draws support an exact result.</p></article>
</div>

<div class="nbd-summary-strip">
<div><span>Current release</span><strong>NBD3 Version 1</strong><small>`r config$labels$project$analysis_period`</small></div>
<div><span>Model series available</span><strong>`r length(model_names_available)`</strong><small>`r paste(model_names_available, collapse = ", ")`</small></div>
<div><span>Uncertainty propagation</span><strong>`r uncertainty_draw_label`</strong><small>Exact intervals are shown only where the draw outputs support the report cell</small></div>
</div>

<div class="report-callout callout-warning">
<div class="callout-title">Current uncertainty status</div>
<div class="callout-body">Version 1 uses one joint uncertainty model. For general completeness, each province receives its own mean-one factor estimated from death-weighted annual aggregate S2 variation across time after age and sex are collapsed. For injuries, the supplied IMS 2009 and later-cleaned FAMHIS 2017 records, eligibility rules, survey weights, mortuary/facility identifiers, and strata determine the national level, provincial profile, cause composition, and their joint sampling uncertainty. NIMS 2000 supplies count-based cause-composition uncertainty. The IMS correction is held through 2009, the IMS and FAMHIS log components are interpolated through 2017, and FAMHIS is held thereafter. HIV/AIDS parameters are sampled from the Stage 05 variance-covariance matrices. Garbage redistribution keeps every expert-approved target but draws a continuous positive target-weight vector for each multi-target rule. Province, South Africa, and national population-group Person estimates are retained within each draw. The report summarises `r uncertainty_draw_label`.</div>
</div>

<div class="historical-source-note"><strong>Historical basis.</strong> The study lineage summarises the initial South African NBD estimates for 2000 (Bradshaw et al., 2003), the Second National Burden of Disease cause-of-death profile for 1997–2012 (Msemburi et al., 2016), and the documented SA NBD cause-list mappings (Awotiwon et al., 2017).</div>

# Contents {#contents}

<div class="contents-intro">The detailed floating contents panel remains available on the left. The compact guide below follows the order of the original visualization storyboard.</div>

<div class="journey-grid report-contents-grid">
<a class="journey-card" href="#hiv-methods"><span>01</span><strong>HIV/AIDS methods</strong><small>NBD2 foundations, transition to NBD3, and the Version 1 R implementation</small></a>
<a class="journey-card" href="#hiv-model-comparison"><span>02</span><strong>HIV/AIDS model comparison</strong><small>NBD2, NBD3-Stata, NBD3-R, THEMBISA, and GBD2023 together</small></a>
<a class="journey-card" href="#injury-methods"><span>03</span><strong>Injury methods</strong><small>IMS/FAMHIS envelope calibration, survey-informed cause interpolation, and direct comparisons</small></a>
<a class="journey-card" href="#injury-model-comparison"><span>04</span><strong>Injury model comparison</strong><small>NBD2, NBD3-Stata, NBD3-R, and GBD2023 together</small></a>
<a class="journey-card" href="#provincial-natural"><span>05</span><strong>Provincial natural deaths</strong><small>Final R estimates with exact uncertainty intervals where supported</small></a>
<a class="journey-card" href="#provincial-injury"><span>06</span><strong>Provincial injury deaths</strong><small>Final R injury causes and broad injury groups</small></a>
<a class="journey-card" href="#population-natural"><span>07</span><strong>Population-group natural deaths</strong><small>Final R population-group trends</small></a>
<a class="journey-card" href="#population-injury"><span>08</span><strong>Population-group injury deaths</strong><small>Final R population-group injury trends with joint uncertainty</small></a>
<a class="journey-card" href="#uncertainty-methods"><span>09</span><strong>Uncertainty methods</strong><small>Joint draws, evidence sources, propagation, interpretation, and current limitations</small></a>
</div>

# 1. HIV/AIDS methods {#hiv-methods}

<div class="section-lead">The HIV/AIDS section follows the established analytical storyboard: the NBD2 formulation, the transition to NBD3, the Version 1 NBD3 formulation, and then one comparison panel containing the two NBD3 implementations, the preceding NBD2 series, and external benchmarks.</div>

## NBD2 HIV/AIDS formulation

For NBD2 (1997–2012), let $D(h,x,t)$ denote completeness-adjusted deaths for cause $h$, age $x$, and time $t$, and let $N(x,t)$ denote population counts.

Mortality was related to antenatal HIV prevalence $p(t)$ by age:

$$
\frac{D(x,t)}{N(x,t)} =
\begin{cases}
\alpha(x) + \beta(x)p(t), & x \geq 5, \\
\exp\{\alpha(x) + \beta(x)p(t)\}, & x < 5.
\end{cases}
$$

A counterfactual background trend was estimated from deaths at ages 75–84:

$$
\frac{D(h,x,t)}{N(x,t)} = \mu_h + \gamma_h\log(1+t).
$$

The age-specific baseline was allocated to causes using the 1997 cause shares:

$$
\alpha(x,h)=\alpha(x)\frac{D(h,x,1997)}{D(x,1997)}.
$$

Misclassified AIDS deaths were estimated as the residual above the projected non-HIV background:

$$
\Omega(h,x,t)=D(h,x,t)-
\alpha(x,h)\left(1+\frac{\gamma_h}{\mu_h}\log(1+t)\right)N(x,t).
$$

## Transition from NBD2 to NBD3 Version 1

<div class="two-column-text">
<div class="support-box">
<strong>Core structure retained</strong>
<ul>
<li>Completeness-adjusted deaths by cause, age, and sex.</li>
<li>Antenatal prevalence as the HIV exposure measure.</li>
<li>Background mortality estimated from older ages.</li>
<li>Cause-level allocation from the 1997 baseline distribution.</li>
<li>Misclassified AIDS deaths treated as an accounting transfer, preserving total deaths.</li>
</ul>
</div>
<div class="boundary-box">
<strong>NBD3 Version 1 refinements</strong>
<ul>
<li>$\gamma_h$ and $\mu_h$ vary by province and sex.</li>
<li>Counterfactual trends are fitted with Poisson models rather than OLS.</li>
<li>Triangular age weights replace a flat step function.</li>
<li>A combined-sex model is used below age five to stabilise sex-specific intercepts.</li>
<li>No misclassified AIDS deaths are estimated at ages 75 years and older.</li>
</ul>
</div>
</div>

## NBD3 Version 1 HIV/AIDS formulation

For 1997–2019, province- and sex-specific counterfactual trends are fitted at ages 75–84:

$$
D(h,x,t)\sim\operatorname{Poisson}\{\lambda(h,x,t)\},\qquad
\log\lambda=\log N+\mu_h+\gamma_h\log(1+t).
$$

The age-specific baseline remains:

$$
\alpha(x,h)=\alpha(x)\frac{D(h,x,1997)}{D(x,1997)}.
$$

The time trend is applied with the triangular age weight:

$$
w(x)=
\begin{cases}
0, & x<1\text{ or }70\leq x<75,\\
0.25, & x\in[1,4]\text{ or }x\in[65,69],\\
0.50, & x\in[5,9]\text{ or }x\in[60,64],\\
0.75, & x\in[10,14]\text{ or }x\in[55,59],\\
1, & x\in[15,54].
\end{cases}
$$

The projected non-HIV death rate and reallocated deaths are:

$$
R_{\mathrm{non-HIV}}(h,x,t)=
\exp\{\alpha(x,h)+w(x)\gamma_h\log(1+t)\},
$$

$$
\Omega(h,x,t)=D(h,x,t)-R_{\mathrm{non-HIV}}(h,x,t)N(x,t).
$$

The transfer is performed within province × sex × year × age × population-group cells, so the all-cause mortality envelope is preserved.

## HIV/AIDS estimates across models {#hiv-model-comparison}

<div class="section-lead">NBD3-Stata and NBD3-R represent the unpublished benchmark implementation and the reproducible R implementation of the same NBD3 Version 1 analysis. They are shown together with NBD2, THEMBISA, and GBD2023 so that replication and external comparison can be assessed in one place.</div>

```{r register-hiv-comparison, include=FALSE}
register_comparison_panel("hiv_compare", "hiv", "hiv_deaths")
```

```{r hiv-comparison-ui}
comparison_panel_ui("hiv_compare", "hiv", "hiv_deaths")
```

# 2. Injury methods {#injury-methods}

<div class="section-lead">The injury analysis has three linked parts: the national total injury level, its province–sex–age profile, and the division of those injury deaths among 15 causes. All IMS 2009 and later-cleaned FAMHIS 2017 injury records meeting the published analytic criteria determine the injury level and profile. Cause fractions use only records mapped to a well-specified common injury cause; generic and unresolved injury records remain in the completeness envelope but are excluded from the cause-fraction denominator. NIMS 2000, IMS 2009, and FAMHIS 2017 then define the cause-composition trajectory.</div>

## Injury cause-fraction data sources

Three mortality surveys provide weighted evidence for the 15 final injury causes:

<div class="anchor-grid">
<div class="anchor-card"><span class="anchor-year">2000</span><h3>NIMS</h3><p>National injury composition by sex, age, and cause. IMS 2009 supplies the province × population-group allocation pattern used to place the national NIMS composition on the complete analytical grid.</p></div>
<div class="anchor-card"><span class="anchor-year">2009</span><h3>IMS</h3><p>Province × population group × sex × age × cause observations, used both as a survey input and as the spatial donor for NIMS.</p></div>
<div class="anchor-card"><span class="anchor-year">2017</span><h3>FAMHIS</h3><p>Later forensic mortality observations harmonised to the common analytical cause set and demographic grid.</p></div>
</div>

```{r injury-source-table}
if (nrow(injury_source_table)) {
  nbd_kable_html(
    injury_source_table,
    caption = paste(
      "Observed weighted injury deaths by source.",
      "Each survey cell gives the weighted deaths and its share of that survey's injury total."
    ),
    align = c("r", "l", rep("r", ncol(injury_source_table) - 2L)),
    escape = FALSE,
    table_class = "nbd-kable injury-survey-table",
    wrap_class = "nbd-kable-wrap injury-survey-table-wrap"
  )
}
```

## Fast survey-informed cause-fraction trajectory

The 15 injury fractions form a closed composition, so the annual trajectory is constructed in a hierarchical additive-log-ratio space rather than by smoothing each fraction independently. Four broad groups are represented first—transport injuries, other unintentional injuries, self-harm, and interpersonal violence—followed by the detailed causes within each broad group.

For survey observation $s$, demographic stratum $j$, non-reference component $h$, and reference component $r$:

$$
z_{hjs}=
\log\left(\frac{p_{hjs}}{p_{rjs}}\right).
$$

The annual path is calculated in five transparent steps:

1. derive the hierarchical log ratios from NIMS 2000, IMS 2009, and FAMHIS 2017;
2. linearly interpolate from 2000 to 2009 and from 2009 to 2017;
3. hold the raw log-ratio path flat before 2000 and after 2017;
4. apply a centred five-year triangular moving average with weights $1,2,3,2,1$, equivalent to applying a centred three-year moving average twice; and
5. transform broad and within-group log ratios back through hierarchical softmax functions.

The fixed moving average spreads survey-year changes over neighbouring years. The survey-year estimates are **not** inserted again after smoothing, so the final series does not jump back to an exact survey fraction. The complete calculation is matrix based and does not fit iterative GAMs.

<div class="report-callout callout-info">
<div class="callout-title">Survey observations are evidence, not exact constraints</div>
<div class="callout-body">The survey systems do not define every detailed injury cause identically. Before interpolation, the production anchors harmonise causes 132 and 138 at NIMS 2000 by backward extrapolation from IMS and FAMHIS, and cause 136 at FAMHIS 2017 by forward extrapolation from NIMS and IMS. The corrected composition is renormalised, after which the same hierarchical interpolation and moving-average rule is applied to all causes. This prevents a change in survey coding from appearing as a steep epidemiological trend.</div>
</div>


## Total injury level and relative profile

The injury model separates the national level from the relative demographic profile. The point anchors are calculated directly from the supplied microdata: IMS retains records with `Cause_of_death == 1` and uses the supplied survey weight; FAMHIS retains records with a nonblank injury mechanism that are not explicitly excluded by `non_nat_undert == 0`, using the supplied `wht`. The published national estimates and intervals remain external validation references and do not replace the empirical totals.

```{r injury-level-reference-audit}
if (nrow(injury_level_audit)) {
  level_display <- injury_level_audit[, .(
    Survey = survey,
    Year = survey_year,
    `Empirical weighted total` = round(empirical_level_estimate),
    `External reference` = round(reference_estimate),
    `Reference lower` = round(reference_lower),
    `Reference upper` = round(reference_upper),
    `Difference` = round(difference, 1),
    `Difference (%)` = round(100 * relative_difference, 3),
    `Well-specified cause mass (%)` = round(100 * well_specified_share, 2),
    `Reference version` = reference_version
  )]
  nbd_kable_html(
    level_display,
    caption = paste(
      "Survey-derived national injury totals compared with external published",
      "benchmarks. The empirical totals, not the external references, define",
      "the NBD3-R injury-level anchors."
    ),
    align = c("l", "r", "r", "r", "r", "r", "r", "r", "r", "l")
  )
}
```

At survey year $y$, the empirical national injury total $S_y$ is compared with the completed routine injury total $V_y$:

$$
K_y=\frac{S_y}{V_y}.
$$

The IMS calibration is held through 2009, the log calibration is linearly interpolated between 2009 and 2017, and the FAMHIS calibration is held through 2019. Separately, the survey and routine injury totals are normalised across province × sex × age, and their share ratio defines the relative profile:

$$
R_{psay}=\frac{q^{S}_{psay}}{q^{R}_{psay}}.
$$

The log relative profile is smoothed over age and follows the same hold/interpolation policy. NIMS remains a cause-composition source rather than an injury-level survey. The earlier assumption that injury completeness changed in proportion to all-cause completeness is not used.

Within every province × sex × population-group × age × year cell, injury and natural deaths remain complementary parts of the completed total. One annual bounded log-odds intercept makes the adjusted demographic profile sum to the survey-calibrated national target. This preserves every completed all-cause cell, prevents negative natural deaths, and prevents injuries from exceeding total deaths.

<div class="report-callout callout-info">
<div class="callout-title">How injury uncertainty is propagated</div>
<div class="callout-body">Each joint draw resamples sampled mortuaries or forensic pathology facilities within the original survey strata. One IMS replicate and one FAMHIS replicate jointly determine the national injury level, provincial and demographic profile, and well-specified cause counts. Thus level, geography, age, sex, and cause composition retain their within-survey covariance. NIMS 2000 uses count-based composition uncertainty because its complete sampling design is not available. The same temporal interpolation, harmonisation, and bounded all-cause calibration are rerun inside every draw.</div>
</div>

```{r injury-composition-chart}
output$injury_composition_chart <- highcharter::renderHighchart({
  if (!nrow(injury_broad$model)) {
    return(empty_nbd_chart(
      "The injury model checkpoints are unavailable.",
      height = 500
    ))
  }
  hc_injury_composition(
    injury_broad$source,
    injury_broad$model,
    config,
    title = "National injury composition: surveys and smoothed interpolation",
    subtitle = paste(
      "Survey diamonds and the smoothed lines use the same national",
      "injury-envelope weighting basis; the survey years are not forced back to the points."
    ),
    height = 500
  )
})
highcharter::highchartOutput("injury_composition_chart", height = "500px")
```

### Survey observations compared with the final trajectory

The table below quantifies the comparison at the four broad injury-group level. A non-zero difference is expected because the same five-year smoother is applied through the survey years rather than resetting the estimates to the observed points.

```{r injury-broad-survey-model-table}
if (nrow(injury_broad_fit_table)) {
  injury_broad_display <- data.table::copy(injury_broad_fit_table)
  numeric_columns <- names(injury_broad_display)[vapply(
    injury_broad_display, is.numeric, logical(1)
  )]
  injury_broad_display[, (numeric_columns) := lapply(.SD, round, 2),
                       .SDcols = numeric_columns]
  nbd_kable_html(
    injury_broad_display,
    caption = paste(
      "Broad injury composition in each survey year compared with the",
      "final smoothed interpolation on the injury-envelope weighting basis."
    )
  )
}
```

## Cause comparability and sparse survey cells

Three detailed causes have known comparability concerns across the survey instruments. The production trajectory therefore uses the documented harmonisation formulas for causes 132, 136, and 138 before the survey compositions are renormalised and interpolated. Both the uncorrected survey fractions and the harmonised inputs are retained in the Stage 03 audit tables.

<div class="interpret-grid">
<div class="support-box"><strong>Sparse cells</strong><p>Where a complete survey stratum has no injury deaths, its composition is borrowed from the most specific observed donor available: province × sex × age, national sex × age, national age, and finally the national composition. Cells sharing the same donor also share the same survey-composition draw in the uncertainty analysis.</p></div>
<div class="boundary-box"><strong>Trajectory interpretation</strong><p>The final line is a deterministic smoothed interpolation of the three survey inputs. Differences between a survey point and the final estimate are shown directly and reflect the common smoothing rule rather than hidden calibration.</p></div>
</div>

```{r injury-flagged-survey-model-table}
if (nrow(injury_flagged_fit_table)) {
  injury_flagged_display <- data.table::copy(injury_flagged_fit_table)
  numeric_columns <- names(injury_flagged_display)[vapply(
    injury_flagged_display, is.numeric, logical(1)
  )]
  injury_flagged_display[, (numeric_columns) := lapply(.SD, round, 3),
                         .SDcols = numeric_columns]
  nbd_kable_html(
    injury_flagged_display,
    caption = paste(
      "Survey inputs and final smoothed estimates for the three detailed",
      "causes with known cross-survey comparability concerns."
    )
  )
}
```

## Injury estimates across models {#injury-model-comparison}

<div class="section-lead">The Stata and R implementations of NBD3 Version 1 are compared directly with the available NBD2 and GBD2023 injury series. GBD uncertainty limits are retained where supplied; exact matching R-implementation cells use the completed joint draws.</div>

```{r register-injury-comparison, include=FALSE}
register_comparison_panel("injury_compare", "injury", "injuries_deaths")
```

```{r injury-comparison-ui}
comparison_panel_ui("injury_compare", "injury", "injuries_deaths")
```

# 3. Final NBD3 Version 1 cause estimates {#final-results}

<div class="section-lead">The four final-results panels follow the established visualization order and present the NBD3 Version 1 estimates from the R implementation. The Stata implementation remains in the comparison panels as the replication benchmark. Shaded intervals appear only where the completed compact uncertainty outputs reproduce the selected cause, age, sex, and geography exactly.</div>

<div class="report-callout callout-info">
<div class="callout-title">What the uncertainty bands cover</div>
<div class="callout-body">The joint-draw output reports Person estimates for provinces, South Africa, and the four national population groups at &lt;1 year, under 5 years, and all ages. It retains all 214 Stage 06 analysis causes, together with the total and injury envelopes. This allows changes in HIV/AIDS to appear as corresponding changes in donor causes and allows the expert-target redistribution draws to propagate into every mapped final ZA cause; the legal-intervention structural-zero row remains a deterministic zero. Crude-rate intervals use the deterministic population denominator. Sex-specific, ASR, YLL, and other age selections remain point estimates because those dimensions or measures are not stored in the current draw profile.</div>
</div>

## Provincial natural deaths {#provincial-natural}

```{r register-provincial-natural, include=FALSE}
register_final_panel("provincial_natural", c("province", "national"), "natural", "za_2")
```

```{r provincial-natural-ui}
final_panel_ui("provincial_natural", c("province", "national"), "natural", "za_2")
```

## Provincial injury deaths {#provincial-injury}

```{r register-provincial-injury, include=FALSE}
register_final_panel("provincial_injury", c("province", "national"), "injury", "za_171")
```

```{r provincial-injury-ui}
final_panel_ui("provincial_injury", c("province", "national"), "injury", "za_171")
```

## Population-group natural deaths {#population-natural}

```{r register-population-natural, include=FALSE}
register_final_panel("population_natural", "population_group", "natural", "za_2")
```

```{r population-natural-ui}
final_panel_ui("population_natural", "population_group", "natural", "za_2")
```

## Population-group injury deaths {#population-injury}

```{r register-population-injury, include=FALSE}
register_final_panel("population_injury", "population_group", "injury", "za_171")
```

```{r population-injury-ui}
final_panel_ui("population_injury", "population_group", "injury", "za_171")
```

# 4. Methods used to derive uncertainty {#uncertainty-methods}

<div class="section-lead">Uncertainty is propagated through one joint simulation rather than calculated separately after the point estimates. Each draw begins with the validated Stage 04 mortality data, varies the evidence-supported components, reruns the injury, HIV/AIDS, and redistribution operations in their analytical order, and then constructs the same final cause hierarchy used for the deterministic database.</div>

## Joint simulation framework

For draw $b$, the complete final cause vector is denoted by $\mathbf{D}^{(b)}$. The reporting interval for a selected result is obtained from the empirical distribution across the completed joint draws:

$$
\left[
Q_{0.025}\left\{D^{(b)}\right\},
Q_{0.975}\left\{D^{(b)}\right\}
\right].
$$

The draws are carried through as complete cause vectors. Therefore, aggregation to all causes, broad causes, final ZA causes, provinces, South Africa, or population groups preserves the covariance generated by the analytical reallocations. The current report uses `r uncertainty_draw_label`; the draw count is controlled in config/uncertainty_joint.yml.

<div class="method-flow uncertainty-flow">
<article><span>1</span><h3>Completeness</h3><p>Draw one province-specific African natural-cause completeness factor for each province.</p></article>
<article><span>2</span><h3>Injuries</h3><p>Draw the IMS/FAMHIS injury envelope and the three survey compositions, then rerun both annual trajectories.</p></article>
<article><span>3</span><h3>HIV/AIDS</h3><p>Draw fitted coefficients jointly and rerun the complete Stage 05 reallocation.</p></article>
<article><span>4</span><h3>Redistribution</h3><p>Draw one continuous positive multiplier vector across the approved targets of each multi-target rule and rerun Stage 06.</p></article>
<article><span>5</span><h3>Aggregation</h3><p>Construct final causes and summarise the within-draw totals by geography and age.</p></article>
</div>

## Evidence and stochastic operations

```{r uncertainty-drivers}
uncertainty_drivers <- data.table::data.table(
  Driver = c(
    "African natural-cause completeness",
    "National injury level and relative profile",
    "Injury cause composition",
    "HIV/AIDS statistical model",
    "Garbage and ill-defined redistribution"
  ),
  `Evidence used` = c(
    paste(
      "Within each province, annual aggregate S2 variation across time",
      "after sex and age are collapsed using implied pre-adjustment deaths"
    ),
    paste(
      "Published-compatible IMS 2009 and later-cleaned FAMHIS 2017 records,",
      "weights, sampled facilities, strata, and the completed routine injury envelope"
    ),
    paste(
      "Well-specified NIMS, IMS and FAMHIS mechanisms; causes 132, 136 and",
      "138 harmonised before interpolation"
    ),
    paste(
      "Fitted Stage 05 coefficient vectors and their estimated",
      "variance-covariance matrices"
    ),
    paste(
      "The original expert-approved target lists and the structural",
      "uncertainty scale of the former non-empty-subset model"
    )
  ),
  `Operation in each draw` = c(
    paste(
      "Use the province-specific log SD to draw one independent",
      "mean-one factor for each province"
    ),
    paste(
      "Select one stratified PSU bootstrap replicate from each survey and",
      "rebuild the level and relative profile jointly"
    ),
    paste(
      "Use the same survey replicate for well-specified IMS/FAMHIS cause",
      "counts; draw NIMS counts and rerun harmonisation and ALR interpolation"
    ),
    paste(
      "Draw coefficients jointly from the fitted covariance matrices",
      "after completeness scaling and rerun Stage 05"
    ),
    paste(
      "Draw positive mean-one Gamma multipliers across all approved targets",
      "and rerun the unchanged ordered proportional redistribution algorithm"
    )
  ),
  `What changes` = c(
    "The African natural mortality envelope, including HIV/AIDS",
    paste(
      "Total injuries and the natural-versus-injury split; the completed",
      "all-cause total is preserved"
    ),
    "The division of the sampled injury envelope among 15 injury causes",
    "The division of deaths between HIV/AIDS and linked non-HIV causes",
    "The recipient cause of redistributed deaths, not the cell total"
  )
)
nbd_kable_html(
  uncertainty_drivers,
  caption = "Evidence and operations in the joint uncertainty model",
  align = c("l", "l", "l", "l")
)
```

### Completeness and under-reporting

Completeness uncertainty is estimated separately for each province. Within each province-year, sex and age are collapsed using implied pre-adjustment African natural deaths as weights, and the weighted standard deviation of the resulting annual aggregate log S2 is calculated across time. One mean-one lognormal factor is then drawn independently for each province and applied to all African natural-cause deaths in that province before HIV/AIDS is re-estimated. This measures variation in the province-wide reporting level over time and does not treat systematic age differences as uncertainty in a common province factor.

The factor is common to every African natural cause within a province. HIV/AIDS therefore receives both province-specific under-reporting uncertainty and uncertainty from the fitted HIV/AIDS model. Population-group estimates retain the complete joint draw at national population-group level, so the African series reflects the completeness draws directly and all four population groups retain the subsequent HIV/AIDS and redistribution covariance.

### Injury level, relative profile and cause composition

IMS 2009 and FAMHIS 2017 determine uncertainty in the national injury level and the relative province–sex–age profile directly from the supplied survey designs. Sampled mortuaries or forensic pathology facilities are resampled with replacement within their survey strata; singleton sampled strata are held fixed. The same replicate simultaneously generates the national total and all provincial and demographic totals, so provinces always sum to the national estimate and their covariance is retained.

The replicate-specific level and profile are compared with the completed routine injury envelope. The IMS correction is held through 2009, the IMS and FAMHIS log components are interpolated from 2009 to 2017, and the FAMHIS correction is held thereafter. Injury and natural deaths remain bounded complements of the same completed all-cause total.

Cause fractions use only well-specified mechanisms on the common 15-cause list. Generic and unresolved survey injuries contribute to total injury completeness but not to the cause-fraction denominator. For IMS and FAMHIS, the same PSU replicate used for the level and profile supplies the well-specified cause counts. NIMS 2000 uses count-based uncertainty. Causes 132, 136, and 138 are harmonised within every draw before the hierarchical additive-log-ratio interpolation and five-year triangular moving average are applied.

<div class="report-callout callout-info">
<div class="callout-title">How to read the injury intervals</div>
<div class="callout-body"><strong>Total injuries retain survey-level uncertainty; provincial and demographic totals retain the joint level/profile uncertainty; specific causes additionally retain cause-composition uncertainty.</strong> The reported IMS and FAMHIS publication estimates remain validation benchmarks. The uncertainty shown by NBD3-R comes from the resampled survey records and design variables rather than being imposed from published confidence limits.</div>
</div>

### HIV/AIDS model uncertainty

For each fitted Stage 05 model, the coefficient vector is sampled jointly:

$$
\boldsymbol{\beta}^{(b)}
\sim
N\left(
\widehat{\boldsymbol{\beta}},
\widehat{\operatorname{Var}}
\left(\widehat{\boldsymbol{\beta}}\right)
\right).
$$

The complete HIV/AIDS calculation is rerun after completeness and injury-envelope scaling. More deaths allocated to HIV/AIDS imply fewer deaths retained by its donor causes in the same demographic cell; fewer HIV/AIDS deaths imply the converse. This creates negative covariance between HIV/AIDS and linked non-HIV causes while preserving the cell total. Antenatal HIV prevalence remains fixed because the supplied prevalence series does not contain an approved variance-covariance source.

### Garbage and ill-defined redistribution

For a rule with approved targets $\{a,b,c\}$, the deterministic point estimate uses the equal multiplier vector $(1,1,1)$. In an uncertainty draw, the same approved targets remain available but receive one continuous positive multiplier vector:

$$
(W_a,W_b,W_c), \qquad W_j>0, \qquad \overline W=1.
$$

If the current target counts are $(D_a,D_b,D_c)$ and the source contains $S$ deaths, the amount added to target $j$ is

$$
A_j=S\frac{D_jW_j}{\sum_k D_kW_k}.
$$

When all current target counts are zero, the normalised multipliers themselves provide the fallback shares. The multipliers are independent Gamma draws before normalisation. Their shape is derived so that, for equal baseline targets, the marginal variance of the continuous allocation share matches the variance under the former uniformly sampled non-empty target-subset model. This preserves the established structural uncertainty scale while removing exact target switching and retaining all expert-approved causes in every draw. Single-target rules remain deterministic.

## Interpretation by result type

```{r uncertainty-by-result}
uncertainty_by_result <- data.table::data.table(
  Result = c(
    "All causes",
    "HIV/AIDS",
    "Other natural causes",
    "Total injuries",
    "Specific injury causes",
    "Population-group causes"
  ),
  `Uncertainty carried into the result` = c(
    paste(
      "General completeness changes the total. Injury-envelope calibration,",
      "HIV/AIDS reallocation, composition, and redistribution preserve it."
    ),
    paste(
      "Completeness, fitted HIV/AIDS coefficients, and any subsequent",
      "eligible redistribution."
    ),
    paste(
      "Completeness, the reciprocal injury-envelope offset in African",
      "natural deaths, HIV/AIDS reallocation, and expert redistribution."
    ),
    paste(
      "IMS/FAMHIS envelope draws remain in the total; survey cause",
      "composition cancels when all injury causes are summed."
    ),
    paste(
      "IMS/FAMHIS envelope draws plus NIMS/IMS/FAMHIS composition",
      "draws, interpolation, smoothing, and eligible redistribution."
    ),
    paste(
      "The same complete joint draw, aggregated nationally within each",
      "population group; no proportional approximation is used."
    )
  ),
  `Changes the all-cause total?` = c(
    "Yes — through completeness",
    "No — HIV/AIDS reallocation preserves the cell total",
    "Completeness changes the total; the injury offset itself preserves it",
    "No — the injury-envelope adjustment reallocates within the completed total",
    "No — the injury-envelope adjustment reallocates within the completed total",
    paste(
      "Yes within a population group: the injury-envelope offset is carried",
      "by African natural deaths, while the provincial all-cause total is preserved"
    )
  )
)
nbd_kable_html(
  uncertainty_by_result,
  caption = "How the joint uncertainty components appear in reported outcomes",
  align = c("l", "l", "l")
)
```

## Geography, aggregation, and covariance

The province and South Africa intervals are generated from one set of per-draw files. National population-group intervals are retained in a separate compact per-draw table so the final hierarchy is constructed separately within each population group. The final report maps the complete analysis-cause vector to the final ZA hierarchy within each draw and only then calculates quantiles. It does not add independent cause-specific variances or allocate provincial intervals proportionally to population groups.

All-cause mortality is recalculated within every draw:

$$
T^{(b)}=\sum_c D_c^{(b)}.
$$

Consequently,

$$
\operatorname{Var}(T)=
\sum_c\operatorname{Var}(D_c)+
2\sum_{c<k}\operatorname{Cov}(D_c,D_k).
$$

The covariance term explains why a cause-specific interval can be visible even when the corresponding all-cause total changes little: reallocation increases one cause while decreasing another.

## Current scope and limitations

```{r uncertainty-limitations}
uncertainty_limitations <- data.table::data.table(
  Quantity = c(
    "Monte Carlo draw count",
    "S1 completeness",
    "Cross-province completeness covariance",
    "Independent NPR variance",
    "Antenatal HIV prevalence",
    "Injury level, profile and composition evidence",
    "Reported measures"
  ),
  `Current treatment` = c(
    paste(
      uncertainty_draw_label,
      "in the configured joint run; increase the draw count for more stable interval tails"
    ),
    "Fixed because no defensible sampling variance is supplied",
    "Not supplied; province-specific completeness factors are drawn independently",
    "Fixed because no independent NPR-specific variance source is supplied",
    "Fixed because no approved prevalence variance-covariance input is supplied",
    paste(
      "IMS/FAMHIS level, provincial profile and well-specified causes use a",
      "stratified PSU bootstrap from the supplied facility and stratum fields;",
      "finite-population corrections and the original software's exact",
      "variance settings are not imposed"
    ),
    paste(
      "Intervals are generated for deaths and corresponding crude rates;",
      "ASRs and YLLs remain point estimates in the current draw profile"
    )
  )
)
nbd_kable_html(
  uncertainty_limitations,
  caption = "Current boundaries of the Version 1 uncertainty analysis",
  align = c("l", "l")
)
```

<div class="report-callout callout-warning">
<div class="callout-title">Interpretation of the current intervals</div>
<div class="callout-body">The current intervals are empirical estimates from `r uncertainty_draw_label`. They represent the uncertainty sources for which the pipeline has an explicit evidence-based stochastic operation. Quantities without a defensible variance source remain fixed rather than being assigned an arbitrary standard deviation. Increasing the draw count improves Monte Carlo stability but does not add uncertainty for quantities that remain fixed.</div>
</div>

<div class="report-footer"><div><strong>Third South African National Burden of Disease Study</strong><br/>Version 1: Reproducible mortality estimates for South Africa, 1997–2019</div><div>`r config$labels$project$analysis_period`</div></div>
